---
title: "Response to Muddy Points"
subtitle: "Lecture: Power and Sample Size for Proportions (Week 9, Day 1)"
author: "Emile"
date: "2026-03-04"
date-format: long
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    html-math-method: mathjax
    embed-resources: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

library(lamisc)
library(laviz)

theme_set(laviz::theme_minimal_white(grid = "none", axis = FALSE, base_size = 14))
```

# Overview

Thank you again for the great feedback! The pace looks excellent — about 91% of you rated it "about right," which is right in the sweet spot. Several of you highlighted the R coding walkthrough and the power analysis for two independent proportions as particularly clear, and it's great to hear that the examples showing the tradeoffs among the four components are connecting to your own research thinking.

::: {.callout-note}
You do not need to read every section below. Each heading addresses a specific muddy point raised in class. Feel free to focus on the ones most relevant to you.
:::

---

# What are concordant and discordant pairs?

> *"I don't understand the concept of the concordant and discordant pairs. Is this something that just came out of the simulated data? How do we determine these proportions?"*

Great question — and this is not specific to the example data. Concordant and discordant pairs are a **general concept** that applies any time you have **paired binary outcomes** — for instance, the same person measured before and after a treatment.

**The idea:** When each person (or matched pair) has two binary outcomes, there are four possible combinations:

::: {style="width: 60%; margin: 0 auto;"}
|  | **After: +** | **After: −** |
|---|:---:|:---:|
| **Before: +** | *a* (concordant) | *b* (discordant) |
| **Before: −** | *c* (discordant) | *d* (concordant) |
:::

- **Concordant pairs** (*a* and *d*): the outcome didn't change. The person was positive both times, or negative both times. These pairs give us **no information about whether the treatment worked** — they tell us nothing about change.

- **Discordant pairs** (*b* and *c*): the outcome **did** change. These are the interesting ones:
  - *b*: changed from + to − (e.g., "got worse")
  - *c*: changed from − to + (e.g., "improved")

**McNemar's test focuses entirely on the discordant pairs.** It asks: among the people whose outcome changed, is the change symmetric (equally likely in both directions), or is it tilted one way?

**How do we determine these proportions in practice?** You cross-tabulate the before and after outcomes for each person. Every person falls into exactly one of the four cells. You don't need to calculate anything complicated — you just count how many people fall into each cell of the 2×2 table.

Here is a concrete example with 10 patients:

```{r}
#| echo: true

library(janitor)

# 10 patients with before/after outcomes
example <- tibble(
  patient = 1:10,
  before  = c("+", "+", "+", "-", "-", "-", "-", "+", "-", "+"),
  after   = c("+", "-", "+", "+", "+", "-", "-", "+", "+", "-")
)

# Cross-tabulate
example %>%
  tabyl(before, after) %>%
  adorn_title(placement = "combined")
```

From this table you can directly read off:

- **Concordant pairs:** patients who stayed the same (before = after)
- **Discordant pairs:** patients whose outcome changed (before ≠ after)

The key point: these categories come from the **study design** (paired measurements), not from anything special about how the data were generated. Any time you measure the same person twice on a binary outcome, you will get concordant and discordant pairs.

---

# The arcsine transformation and Cohen's *h*

> *"The arcsine stuff was confusing sort of."*

This is understandable — the formula for Cohen's *h* looks unusual:

$$h = 2\arcsin(\sqrt{p_1}) - 2\arcsin(\sqrt{p_2})$$

The good news: **you never need to compute this by hand.** The `ES.h()` function in the `pwr` package does it for you. What matters is understanding **why** it exists.

**The core problem:** A difference of 5 percentage points doesn't always represent the same statistical "difficulty." Consider:

- Going from 50% to 45%: the variance of $\hat{p}$ is high near 0.50, so there is a lot of noise — this difference is hard to detect
- Going from 10% to 5%: the variance is much smaller near the extremes, so there is less noise — this same 5-point difference is easier to detect

```{r}
library(pwr)

# Same absolute difference (0.05), different effect sizes
ES.h(p1 = 0.50, p2 = 0.45)
ES.h(p1 = 0.10, p2 = 0.05)
```

The arcsine transformation accounts for this by "stretching" the scale near 0 and 1 (where proportions are less variable) and "compressing" it near 0.5 (where proportions are most variable). After transformation, equal gaps on the *h* scale correspond to equally detectable differences — regardless of where on the 0-to-1 scale you are.

**The practical takeaway for this course:**

1. You need to specify **two actual proportions** (e.g., $p_0 = 0.30$ and $p_1 = 0.50$), not a single standardized number
2. `ES.h(p1, p2)` converts those two proportions into a single effect size for you
3. You pass that effect size into `pwr.p.test()` or `pwr.2p.test()`
4. You don't need to memorize the arcsine formula — just understand that it corrects for the fact that the same raw difference is not equally detectable at all baseline proportions

Think of it as analogous to Cohen's *d* for means: dividing by $\sigma$ standardizes the difference so it's comparable across different scales. Cohen's *h* does the same thing for proportions, just with a different mathematical tool.

---

# Correlated observations vs. independence

> *"It was a little unclear to me the difference between correlated observations in the context of independence, but that of course makes sense when I put it in the context of how I've typically understood correlations (two observations are related), so that eventually clicked."*

Even though this one eventually clicked, the terminology can be confusing because "correlation" and "independence" get used in overlapping ways across statistics.

Here's the distinction as it applies to choosing the right test:

**Independent observations** means the outcome for one person tells you nothing about the outcome for another person. This is the assumption behind the two-proportion $z$-test (`prop.test()`) and the chi-squared test. If I know Patient A responded to treatment, that gives me no information about whether Patient B responded.

**Correlated (paired) observations** means the two measurements are linked — typically because they come from the **same person** measured twice (before/after), or from matched pairs. If I know a patient failed the screening test before the training, that gives me some information about their result after (maybe they are generally less knowledgeable, or maybe they are exactly the person who benefits most).

The parallel to t-tests might help solidify this:

::: {style="width: 90%; margin: 0 auto;"}
| Design | Means test | Proportions test |
|---|---|---|
| Different people in each group | Two-sample t-test | Two-proportion test / Chi-squared |
| Same person measured twice | Paired t-test | McNemar's test |
:::

Using the wrong test — treating paired data as independent or vice versa — leads to incorrect standard errors and unreliable $p$-values. So the design question ("are these the same people or different people?") always comes before the analysis question.

---

# Pace and general comments

The pace continues to look great — 91% "about right" with one response each at "slightly too slow" and "slightly too fast," which is about as balanced as it gets. Several of you highlighted the R walkthrough and the power analysis examples as particularly clear.

The comment about seeing tradeoffs among the four power components and how that connects to research planning is exactly what I'm hoping for. Power analysis is one of those topics that can feel abstract in the classroom but becomes very concrete when you are writing a grant or designing your own study. 

---

# Summary

- **Concordant vs. discordant pairs:** Concordant pairs (outcome didn't change) give no information about treatment effect; discordant pairs (outcome changed) are what McNemar's test analyzes. You determine them by cross-tabulating the paired outcomes.
- **The arcsine transformation:** Corrects for the fact that proportions near 0.5 are noisier than proportions near 0 or 1. You don't compute it by hand — `ES.h(p1, p2)` handles it. Just remember to specify two actual proportions.
- **Correlated vs. independent observations:** "Same person measured twice" → paired/correlated → McNemar's test. "Different people in each group" → independent → two-proportion test or chi-squared. The design determines the analysis.
