---
title: "Schedule"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    page-layout: full
editor_options: 
  chunk_output_type: console
---

```{css, echo=FALSE}
.title{
  font-size: 30px;
  color: #3070BF;
  background-color: #fff;
  padding: 0px;
}
```

```{r setup, include=FALSE}
library(here)
source(here("class_dates.R"))
```

| Week | Date | Les-son | Topic | TB sections | Key Info | Slides HTML | Slides PDF | Slides Notes | Exit tix | Record-ing | Muddy Points |
|------|------|------|------|------|:----:|:----:|:----:|:----:|:----:|:----:|:----:|
| 1 | `{r} w1d1 %>% format("%m/%d")` | 0 | Welcome | X.X | Key info to come  | HTMl Slides  | PDF slides |  |  |  |
|  |  | 1 | Data collection | 1.3 |Key info to come  | HTMl Slides  | PDF slides |  |  |  |
|  |
|  |  |  |  |  |  |  |  |
|  |  |  |  |  |  |  |  |  |  |  |  |


: {.striped .hover tbl-colwidths="\[6, 6, 6, 46,6, 6, 6, 6, 6,6, 6\]"}




| Week | Date | Lesson | Topic           | TB sections | Key Info         | Slides HTML | Slides PDF | Slides Notes | Exit tix | Recording | Muddy Points |
|------|------|--------|-----------------|-------------|------------------|-------------|------------|--------------|----------|-----------|--------------|
| 1    | `r format(w1d1, "%m/%d")` | 0      | Welcome         | X.X         | Key info to come | HTML Slides | PDF slides  |              |          |           |              |
|      |      | 1      | Data collection | 1.3         | Key info to come | HTML Slides | PDF slides  |              |          |           |              |

: {.striped .hover tbl-colwidths="[6,6,6,46,6,10,8,8,10,6,8,10]"}


| Week | Date | Lesson | Topic           | TB sections | Key Info         | Slides HTML | Slides PDF | Slides Notes | Exit tix | Recording | Muddy Points |
|------|------|--------|-----------------|-------------|------------------|-------------|------------|--------------|----------|-----------|--------------|
| 1    | `r format(w1d1, "%m/%d")` | 0      | Welcome         | X.X         | Key info to come | HTML        | PDF        |              |          |           |              |
|      |      | 1      | Data collection | 1.3         | Key info to come | HTML        | PDF        |              |          |           |              |
|      |      | 2      | Data ethics     | 1.4         | Key info to come | HTML        | PDF        |              |          |           |              |


| Week | Date | Lesson | Topic           | TB sections | Key Info | Slides HTML | Slides PDF | Slides Notes | Exit tix | Recording | Muddy Points |
|------|------|--------|-----------------|-------------|----------|------------|------------|--------------|----------|-----------|--------------|
| **Week 1** | `r format(w1d1, "%m/%d")` |        |                 |             |          |            |            |              |          |           |              |
| 1    |      | 0      | Welcome         | X.X         | ...      | HTML       | PDF        |              |          |           |              |
|      |      | 1      | Data collection | 1.3         | ...      | HTML       | PDF        |              |          |           |              |



```{r}
library(dplyr)
library(gt)

sched <- tibble::tribble(
  ~Week, ~Date, ~Lesson, ~Topic, ~TB, ~KeyInfo,
  1, format(w1d1, "%m/%d"), 0, "Welcome", "X.X", "Key info to come",
  1, format(w1d1, "%m/%d"), 1, "Data collection", "1.3", "Key info to come"
)

sched %>%
  gt(groupname_col = "Week") %>%
  tab_options(row_group.as_column = FALSE)
```

| Week | Date | Lesson | Topic           | TB sections | Key Info | Slides HTML | Slides PDF | Slides Notes | Exit tix | Recording | Muddy Points |
|------|------|--------|-----------------|-------------|----------|------------|------------|--------------|----------|-----------|--------------|
| **Week 1** |  |  |  |  |  |  |  |  |  |  |  |
| 1 | `r format(w1d1, "%m/%d")` | 0 | Welcome |  |  |  |  |  |  |  |  |
|   |  | 1 | Data collection | 1.3 |  |  |  |  |  |  |  |
|   | `r format(w1d2, "%m/%d")` | 2 | Intro to data & numerical summaries | 1.2, 1.4 |  |  |  |  |  |  |  |
|   | **`r format(hw0, "%m/%d")`** |  | **HW 0 due @ 11 pm** |  |  |  |  |  |  |  |  |