---
title: "Correlation and Simple Linear Regression"
subtitle: "Textbook Sections 1.6, 6.1-6.4"
author: "Emile Latour, Nicky Wakim, Meike Niederhausen"
date: "`r library(here); source(here('class_dates.R')); w10d2`"
date-format: long
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v5.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | Correlation and Simple Linear Regression"
    html-math-method: mathjax
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(broom)
library(glue)
library(here)
library(knitr)
library(oibiostat)
library(rstatix)
library(janitor)
library(gt)

library(lamisc)
library(laviz)

theme_set(laviz::theme_minimal_white(grid = "none",
                                     axis = "xy",
                                     base_size = 16))

# Load and clean NHANES data
data(nhanes.samp)

nhanes_bp <- nhanes.samp %>%
  filter(!is.na(BMI), !is.na(BPSys1)) %>%
  select(BMI, BPSys1, Gender, Age)

# Fit the model once — reference throughout
model_bp <- lm(BPSys1 ~ BMI, data = nhanes_bp)
model_tidy <- tidy(model_bp, conf.int = TRUE)
model_glance <- glance(model_bp)

b0 <- lamisc::fmt_num(model_tidy$estimate[1], accuracy = 0.01)
b1 <- lamisc::fmt_num(model_tidy$estimate[2], accuracy = 0.01)
n_bp <- nobs(model_bp)
```

# Learning Objectives

By the end of today's lecture, you will be able to:

1. Calculate and interpret Pearson and Spearman correlation coefficients
2. Identify the components of the simple linear regression model
3. Explain what ordinary least squares (OLS) minimizes
4. Run a simple linear regression in R and interpret the output
5. Check the LINE assumptions using diagnostic plots
6. Interpret $R^2$ as a measure of model fit
7. Conduct and interpret inference for the population slope $\beta_1$
8. Construct confidence and prediction intervals for new observations

## Roadmap for Today

\

::::: columns
::: {.column width="50%"}
**Part 1: Correlation**

- Visualizing two continuous variables
- Pearson correlation: $r$
- Spearman correlation: a nonparametric bridge
- Correlation ≠ causation

**Part 2: The SLR Model**

- From correlation to regression
- Population model vs. estimated line
- Residuals and OLS
- Running `lm()` in R
- Interpreting slope and intercept
:::

::: {.column width="50%"}
**Part 3: Model Evaluation**

- LINE assumptions
- Diagnostic plots in R
- $R^2$: how much variation does the model explain?

**Part 4: Inference and Prediction**

- Hypothesis test for slope $\beta_1$
- Confidence interval for $\beta_1$
- Predicting new observations
- CI for mean response vs. prediction interval
:::
:::::

# Part 1: Correlation

## The question: do two continuous variables move together?


::: columns
::: {.column width="50%"}
\

We've spent the semester asking questions like:

- Does **mean** BP differ between groups?
- Is **proportion** with hypertension the same across age groups?

\

Now a new type of question: **As BMI increases, does systolic blood pressure tend to increase too?**

\

Both variables are continuous. We need new tools.
:::
::: {.column width="50%"}

\

\

\

\

\

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| echo: false

ggplot(nhanes_bp, aes(x = BMI, y = BPSys1)) +
  geom_point(alpha = 0.4, color = "#0072B2", size = 1.5) +
  labs(title = "Systolic blood pressure vs. BMI",
       subtitle = glue("NHANES sample, n = {n_bp}"),
       x = "BMI (kg/m²)",
       y = "Systolic BP (mmHg)")
```
:::
:::


## Response and explanatory variables


::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Response variable

The variable that **measures the outcome of interest** — what we are trying to understand or predict.

**Example:** Systolic blood pressure
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Explanatory variable

The variable we think may **explain or be associated with** changes in the response.

**Example:** BMI
:::

:::
:::::


- A study will typically examine whether the values of a **response variable** differ as values of an **explanatory variable** change
- These roles guide how we set up plots and models
   - The explanatory variable goes on the $x$-axis, the response on the $y$-axis.
- Sometimes we’re just interested in viewing the relationship between **explanatory variables**
- Approaches vary depending on whether the two variables are:
   - Both numerical
   - Both categorical
   - One numerical and one categorical
   
\

When both variables are numerical, the **scatterplot** is our primary tool for visualizing the relationship.
  

## Reading a scatterplot

\

When you look at a scatterplot, ask four questions:

\

::::: columns
::: {.column width="48%"}
**1. Direction**

- Positive: as $X$ increases, $Y$ tends to increase
- Negative: as $X$ increases, $Y$ tends to decrease
- None: no clear pattern

**2. Form**

- Linear: points follow a straight line
- Curved: some other shape
- No pattern

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
**3. Strength**

- How tightly do the points follow the pattern?
- Strong, moderate, weak

**4. Unusual observations**

- Outliers that don't fit the trend
- Clusters or subgroups

:::
:::::

\



<!-- ::: {.callout-tip icon="false"} -->
<!-- For systolic BP vs. BMI: **positive, roughly linear, moderate strength**. Some variability — not a tight cloud. -->
<!-- ::: -->

## Scatterplot for systolic BP vs. BMI

::: columns
::: {.column width="40%"}

1. Direction 
   - Positive
2. Form
   - Roughly linear
3. Strength
   - Moderate
   - Not a tight cloud
4. Unusual observations
   - Some present

:::
::: {.column width="60%"}

\


```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| echo: false

ggplot(nhanes_bp, aes(x = BMI, y = BPSys1)) +
  geom_point(alpha = 0.4, color = "#0072B2", size = 1.5) +
  labs(title = "Systolic blood pressure vs. BMI",
       subtitle = glue("NHANES sample, n = {n_bp}"),
       x = "BMI (kg/m²)",
       y = "Systolic BP (mmHg)")
```
:::
:::

## The Pearson correlation coefficient

\

The **Pearson correlation coefficient** $r$ quantifies the strength and direction of the **linear** association between two continuous variables.

$$r = \frac{1}{n-1}\sum_{i=1}^{n}\left(\frac{x_i - \bar{x}}{s_x}\right)\left(\frac{y_i - \bar{y}}{s_y}\right)$$

\

::::: columns
::: {.column width="48%"}

<!-- **Properties of $r$:** -->

<!-- - Always between −1 and +1 -->
<!-- - $r = +1$: perfect positive linear relationship -->
<!-- - $r = -1$: perfect negative linear relationship -->
<!-- - $r = 0$: no linear relationship -->
<!-- - Sign = direction; magnitude = strength -->

**With base R:**

```{r}
cor(nhanes_bp$BMI, nhanes_bp$BPSys1,
    use = "complete.obs")

```

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**Using tidyverse:**

```{r}
nhanes_bp %>%
  summarise(
    r = cor(BMI, BPSys1,
            use = "complete.obs")
  )
```

:::
:::::

## Describing associations between 2 numerical variables


\

- Two variables $X$ and $Y$ are 

  - **Positively associated** if $Y$ increases as $X$ increases

  - **Negatively associated** if $Y$ decreases as $X$ increases
  
\
- If there is no association between the variables, then we say they are **uncorrelated** or **independent**

\

- The term "association" is a very general term

  - Can be used for numerical or categorical variables
  - Not specifically referring to linear associations

## (Pearson) Correlation coefficient ($r$)

::: columns
::: {.column width="40%"}
* $r = -1$ indicates a __perfect negative linear relationship__: As one variable increases, the value of the other variable tends to go down, following a *straight line*
* $r = 0$ indicates __no linear relationship__: The values of both variables go up/down independently of each other
* $r = 1$ indicates a __perfect positive linear relationship__: As the value of one variable goes up, the value of the other variable tends to go up as well in a linear fashion

 

* The closer $r$ is to ±1, the stronger the linear association
:::

::: {.column width="60%"}

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7
#| fig-align: center

set.seed(620)
n <- 150

# make_scatter <- function(r_target, title) {
#   # Simulate bivariate normal
#   Sigma <- matrix(c(1, r_target, r_target, 1), 2, 2)
#   dat <- as.data.frame(MASS::mvrnorm(n, mu = c(0, 0), Sigma = Sigma))
#   r_obs <- round(cor(dat$V1, dat$V2), 2)
#   ggplot(dat, aes(x = V1, y = V2)) +
#     geom_point(alpha = 0.5, color = "#0072B2", size = 1.2) +
#     geom_smooth(method = "lm", se = FALSE, color = "#D55E00", linewidth = 1) +
#     labs(title = glue("r = {r_obs}"), x = NULL, y = NULL) +
#     theme(axis.text = element_blank(), axis.ticks = element_blank())
# }


make_scatter <- function(r_target, title) {
  # Simulate bivariate normal
  Sigma <- matrix(c(1, r_target, r_target, 1), 2, 2)
  dat <- as.data.frame(MASS::mvrnorm(n, mu = c(0, 0), Sigma = Sigma))
  r_obs <- lamisc::fmt_num(cor(dat$V1, dat$V2), accuracy = 0.01)
  ggplot(dat, aes(x = V1, y = V2)) +
    geom_point(alpha = 0.5, color = "#0072B2", size = 1.8) +
    # geom_smooth(method = "lm", se = FALSE, color = "#D55E00", linewidth = 1) +
    labs(caption = glue("r = {r_obs}"), x = NULL, y = NULL) +
    laviz::theme_minimal_white(grid = "none", 
                               border = TRUE) + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        plot.caption = element_text(
      hjust = 0.5, # Center the caption horizontally
      vjust = 0,   # Align vertically as needed (0 is bottom)
      size = 25,   # Adjust size
      face = "bold" # Adjust font face)
        ))
}

p1 <- make_scatter(0.33)
p2 <- make_scatter(0.69)
p3 <- make_scatter(0.98)
p4 <- make_scatter(1.0)
p5 <- make_scatter(-0.08)
p6 <- make_scatter(-0.64)
p7 <- make_scatter(-0.92)
p8 <- make_scatter(-1.0)

library(patchwork)

p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + 
  patchwork::plot_layout(ncol = 4)

```
:::
:::

## Interpreting the correlation coefficient

Multiple published schemes exist for labeling the strength of $r$ — and they don't agree:

::::: columns
::: {.column width="30%"}

**Common scheme**

| $|r|$ | Label |
|---|---|
| $\geq 0.6$ | Strong |
| $0.3 – 0.6$ | Moderate |
| $< 0.3$ | Weak |

:::

::: {.column width="5%"}
:::

::: {.column width="30%"}

**Another common scheme**

| $|r|$ | Label |
|---|---|
| $> 0.8$ | Strong |
| $0.5 – 0.8$ | Moderate |
| $< 0.5$ | Weak |

:::

::: {.column width="5%"}
:::

::: {.column width="30%"}

**Mukaka (2012)**

| $|r|$ | Label |
|---|---|
| $0.9 – 1.0$ | Very high |
| $0.7 – 0.9$ | High |
| $0.5 – 0.7$ | Moderate |
| $0.3 – 0.5$ | Low |
| $0.0 – 0.3$ | Negligible |

:::
:::::

::: {.callout-important icon="false"}
### The disagreement is the lesson

Three published schemes, three different answers for the same $r$. Rather than memorizing cutoffs, ask: does this correlation **accurately and meaningfully represent the relationship** for your research question and field?

In biomedical research with a single predictor, $r = 0.40$ may be quite meaningful. In a tightly controlled physical measurement, $r = 0.85$ might be disappointing. **Always report the actual value and interpret it in context.**
:::

[Eight things you need to know about interpreting correlations (from University of Sussex)](https://users.sussex.ac.uk/~grahamh/RM1web/Eight%20things%20you%20need%20to%20know%20about%20interpreting%20correlations.pdf)

<!-- ::: {.callout-warning icon="false"} -->
<!-- ### Common rough benchmarks (context-dependent!) -->
<!-- $|r| < 0.3$: weak  &nbsp;&nbsp;&nbsp; $0.3 \leq |r| < 0.6$: moderate  &nbsp;&nbsp;&nbsp; $|r| \geq 0.6$: strong -->

<!-- These are rough rules of thumb — what counts as a "strong" correlation varies by field. In complex biological systems, $r = 0.4$ may be quite meaningful. -->
<!-- ::: -->

<!-- ## Interpreting the correlation coefficient -->

<!-- \ -->

<!-- ::: {.callout-warning icon="false"} -->
<!-- ### There is no universal benchmark -->

<!-- A correlation coefficient should accurately reflect the strength of the relationship in *your* data. What counts as "strong" depends entirely on the subject area. -->
<!-- ::: -->

<!-- \ -->

<!-- ::::: columns -->
<!-- ::: {.column width="48%"} -->

<!-- **One common scheme:** -->

<!-- | Strength | $|r|$ | -->
<!-- |---|---| -->
<!-- | Weak | $< 0.5$ | -->
<!-- | Moderate | $0.5$ to $0.8$ | -->
<!-- | Strong | $> 0.8$ | -->

<!-- ::: -->

<!-- ::: {.column width="4%"} -->
<!-- ::: -->

<!-- ::: {.column width="48%"} -->

<!-- **Why context matters:** -->

<!-- - Studies involving **human behavior or biology** often yield $r < 0.6$ — humans are complex and hard to predict -->
<!-- - Studies of **physical or chemical processes** with precise measurement may yield $r$ near $\pm 1$ -->
<!-- - In biomedical research, $r = 0.4$ with a single predictor can be quite meaningful -->

<!-- ::: -->
<!-- ::::: -->

<!-- \ -->

<!-- ::: {.callout-tip icon="false"} -->
<!-- **Bottom line:** Don't chase a benchmark. Ask whether the correlation accurately and meaningfully represents the relationship for your research question. -->
<!-- ::: -->

## Correlation ≠ causation

\

A high $r$ tells you two variables are associated. It says **nothing** about whether one causes the other.

\

::::: columns
::: {.column width="48%"}

**Classic confounders:**

- Ice cream sales and drowning deaths both increase in summer (confounder: temperature)
- Shoe size and reading ability in children (confounder: age)
- Higher hospital admission rates associated with worse outcomes (confounder: severity of illness)

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**For our example:**

BMI and systolic BP are associated — but the causal pathway is complex:

- BMI → metabolic syndrome → hypertension
- Shared causes: diet, physical activity, genetics
- Reverse causation is unlikely here but not impossible

:::
:::::

\

::: {.callout-important icon="false"}
Correlation is a description of the data. Causation requires study design (randomized experiments) or careful causal reasoning — not a higher $r$.
:::

## Spearman correlation: the nonparametric alternative

\

**Recall from Lesson 17:** when data are skewed, contain outliers, or are ordinal, rank-based methods are more robust.

\

The **Spearman correlation** $r_s$ is simply the Pearson correlation applied to the **ranks** of $X$ and $Y$:

1. Rank all $X$ values (1 = smallest)
2. Rank all $Y$ values (1 = smallest)
3. Compute Pearson $r$ on those ranks

\


**When to use Spearman:**

- Ordinal outcomes (pain scales, severity scores)
- Skewed continuous variables (viral loads, antibody titers)
- When outliers would distort Pearson $r$



## Spearman correlation in R

\

::::: columns
::: {.column width="48%"}

**In base R**

```{r}
cor(nhanes_bp$BMI, nhanes_bp$BPSys1,
    method = "spearman",
    use = "complete.obs")
```


:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**More tidyverse way:**

```{r}
nhanes_bp %>%
  summarise(
    r_pearson = cor(BMI, BPSys1,
        use = "complete.obs"),
    r_spearman = cor(BMI, BPSys1,
        method = "spearman",
        use = "complete.obs")
  )
```

:::
:::::

\

::: {.callout-tip icon="false"}
Pearson and Spearman are similar here — the relationship is reasonably linear without extreme outliers dominating.
:::

## The limitation of correlation

\

Correlation answers: **how strongly are these two variables associated?**

\

But researchers often want to know more:

- How much does systolic BP **increase** for each 1-unit increase in BMI?
- What systolic BP would we **predict** for a patient with BMI = 32?
- Is the association **statistically significant** — or could it be due to chance?

\

::: {.callout-note icon="false"}
### Correlation → Regression

Correlation gives us a single number summarizing the association. **Regression** gives us a line — a model that lets us quantify, predict, and test.

We keep the same scatterplot. We add a line.
:::

# Part 2: The Simple Linear Regression Model

## The regression line

\

A **regression line** describes how the mean of $Y$ changes as $X$ changes:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$$
\

The **hat** (^) on $\hat{Y}$ signals this is an **estimated** value from the line — not the actual observed $Y$.

\

```{r}
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| code-fold: true

ggplot(nhanes_bp, aes(x = BMI, y = BPSys1)) +
  geom_point(alpha = 0.3, color = "#0072B2", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE,
              color = "#D55E00", linewidth = 1.5) +
  labs(title = "Systolic BP vs. BMI with regression line",
       x = "BMI (kg/m²)", y = "Systolic BP (mmHg)")
```



## Population model vs. estimated line

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Population model (unobservable)

$$Y = \beta_0 + \beta_1 X + \epsilon$$

- $\beta_0$, $\beta_1$: unknown population parameters we want to estimate
- $\epsilon$ (epsilon): random error — how far each individual deviates from the true line
- Assumption: $\epsilon \sim N(0, \sigma^2)$ — errors are normally distributed with mean 0 and constant variance
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-tip icon="false"}
### Estimated line (from our data)

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$$

- $\hat{\beta}_0$, $\hat{\beta}_1$: our sample estimates of $\beta_0$, $\beta_1$
- No error term — the line gives the **predicted mean** at each $X$
- Individual observations will scatter above and below this line
:::

:::
:::::

| Symbol | Name | Type |
|---|---|---|
| $Y$ | Response / outcome | Observed |
| $X$ | Predictor / covariate | Observed |
| $\beta_0$ | Intercept | Population parameter |
| $\beta_1$ | Slope | Population parameter |
| $\epsilon$ | Error term | Unobservable |
| $\hat{\beta}_0$, $\hat{\beta}_1$ | Estimated intercept / slope | Calculated from data |

## Residuals: the building block of regression

\

For each observation $i$, the **residual** is the difference between what we observed and what the line predicted:

$$\hat{\epsilon}_i = Y_i - \hat{Y}_i$$

\

::::: columns
::: {.column width="30%"}

\

- Points **above** the line: 
   - positive residual ($Y_i > \hat{Y}_i$)
- Points **below** the line: 
   - negative residual ($Y_i < \hat{Y}_i$)
- Points **on** the line: 
   - residual = 0 (rare in practice)

:::


::: {.column width="70%"}
```{r}
#| fig-width: 9
#| fig-height: 7
#| fig-align: center
#| echo: false

aug <- augment(model_bp)

ggplot(aug, aes(x = BMI, y = BPSys1)) +
  geom_segment(aes(xend = BMI, yend = .fitted),
               color = "#CC79A7", alpha = 0.4, linewidth = 0.5) +
  geom_smooth(method = "lm", se = FALSE,
              color = "#D55E00", linewidth = 0.5) +
  geom_point(alpha = 0.4, color = "#0072B2", size = 1.75) +
  geom_point(aes(y = .fitted), shape = 1,
             color = "#D55E00", size = 1.75) +
  labs(title = "Residuals = observed − fitted",
       subtitle = "Pink segments show residuals; open circles show fitted values",
       x = "BMI (kg/m²)", y = "Systolic BP (mmHg)")
```

:::
:::::

## Ordinary Least Squares (OLS)

\

**Key question:** With infinitely many possible lines we could draw through the data, which one is *best*?

\

**OLS answer:** Choose the line that minimizes the **Sum of Squared Errors (SSE)**:

$$SSE = \sum_{i=1}^{n} \hat{\epsilon}_i^2 = \sum_{i=1}^{n}(Y_i - \hat{Y}_i)^2$$

\

::::: columns
::: {.column width="55%"}

**Why square the residuals?**

- Positive and negative residuals would cancel
- Squaring penalizes large errors more than small ones
- Makes the math tractable (calculus gives closed-form solutions)

The OLS estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ are the values that make SSE as small as possible. R's `lm()` finds these automatically.

- [Math details if interested](https://www.stat.cmu.edu/~hseltman/309/Book/chapter9.pdf)

:::

::: {.column width="5%"}
:::

::: {.column width="40%"}

::: {.callout-tip icon="false"}
### The OLS formulas

$$\hat{\beta}_1 = r \cdot \frac{s_Y}{s_X}$$

$$\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1 \bar{X}$$

Notice: $\hat{\beta}_1$ depends directly on the Pearson correlation $r$ — regression and correlation are deeply connected.
:::

:::
:::::

## Running SLR in R: `lm()` + `tidy()`

\

```{r}
# Fit the model
model_bp <- lm(BPSys1 ~ BMI, data = nhanes_bp)

library(broom)
# Tidy output: coefficients, SE, t, p, CI
tidy(model_bp, conf.int = TRUE)
```

\

Our estimated regression equation:

$$\widehat{\text{Systolic BP}} = `r b0` + `r b1` \times \text{BMI}$$

## Interpreting the slope and intercept

\

$$\widehat{\text{Systolic BP}} = `r b0` + `r b1` \times \text{BMI}$$


::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Slope ($\hat{\beta}_1$ = `r b1`)

For every **1 kg/m² increase in BMI**, systolic blood pressure is expected to increase by **`r b1` mmHg**, on average, holding all else constant.

- Sign tells you the **direction** of the association
- Magnitude tells you the **rate of change** in the units of $Y$ per unit of $X$
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Intercept ($\hat{\beta}_0$ = `r b0`)

The expected systolic BP for someone with a BMI of 0 is `r b0` mmHg.

**⚠️ The intercept is often not directly interpretable.** A BMI of 0 is physiologically impossible. The intercept is a mathematical anchor for the line — interpret it only when $X = 0$ is meaningful in context.
:::

:::
:::::

::: {.callout-important icon="false"}
Regression describes **association**, not causation. "Expected to increase by `r b1` mmHg" does not mean BMI *causes* higher blood pressure — it means these two variables are associated in a positive linear way.
:::

# Part 3: Model Evaluation

## The LINE conditions

\

For the regression model to give valid estimates and reliable inference, four conditions must hold. We remember them as **LINE**:

\

| Letter | Condition | How to check |
|:---:|---|---|
| **L** | **L**inearity — the relationship between $X$ and $Y$ is linear | Scatterplot; residual plot |
| **I** | **I**ndependence — observations are independent of each other | Study design |
| **N** | **N**ormality — residuals are approximately normally distributed | QQ plot of residuals |
| **E** | **E**qual variance — residuals have constant variance across all $X$ values | Residual plot |

: {tbl-colwidths="\[7, 50, 25\]"}

\

::: {.callout-tip icon="false"}
The key diagnostic plots are the **residual plot** (residuals vs. fitted values) and the **QQ plot of residuals**. We generate both from the fitted model object.
:::

## Diagnostic plots in R

```{r}
#| fig-width: 12
#| fig-height: 4.5
#| fig-align: center
#| code-fold: true

# TODO -- pick up here

aug <- augment(model_bp)

p1 <- ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4, color = "#0072B2", size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_smooth(se = FALSE, color = "#D55E00", linewidth = 1) +
  labs(title = "Residual plot",
       subtitle = "Checks L and E",
       x = "Fitted values (ŷ)",
       y = "Residuals")

p2 <- ggplot(aug, aes(sample = .resid)) +
  stat_qq(alpha = 0.4, color = "#0072B2", size = 1.5) +
  stat_qq_line(color = "#D55E00", linewidth = 1) +
  labs(title = "Normal QQ plot of residuals",
       subtitle = "Checks N",
       x = "Theoretical quantiles",
       y = "Sample quantiles")

library(patchwork)
p1 + p2
```

```{r}
model_bp <- lm(BPSys1 ~ BMI, data = nhanes_bp)
```



## Reading the diagnostic plots

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Residual plot (checks L and E)

**What you want to see:**

- Points scattered randomly around the horizontal line at 0
- No systematic curve or funnel shape
- Roughly equal spread across all fitted values

**Red flags:**

- **Curve:** linearity (L) is violated — try a transformation
- **Funnel (spread increases):** equal variance (E) is violated — heteroscedasticity
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### QQ plot (checks N)

**What you want to see:**

- Points fall approximately along the diagonal reference line
- [Understanding QQ Plots, Univeristy of Virginia](https://library.virginia.edu/data/articles/understanding-q-q-plots)

**Red flags:**

- S-shaped pattern: skewed residuals
- Heavy tails: points deviate at both ends

**Note:** With large samples, moderate departures from normality matter less (CLT). Linearity and equal variance are more critical for inference.
:::

:::
:::::

## The $R^2$ statistic

\

**$R^2$** (the coefficient of determination) tells us: **what proportion of the variability in $Y$ is explained by the linear relationship with $X$?**

$$R^2 = 1 - \frac{SSE}{SST} = 1 - \frac{\sum(Y_i - \hat{Y}_i)^2}{\sum(Y_i - \bar{Y})^2}$$

\

::::: columns
::: {.column width="48%"}

- $R^2 = 0$: the model explains none of the variability in $Y$ — no better than predicting $\bar{Y}$ for everyone
- $R^2 = 1$: the model explains all of the variability — all points fall exactly on the line
- For simple linear regression: $R^2 = r^2$ (square of the Pearson correlation)

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

```{r}
# glance() gives model-level summaries
glance(model_bp) %>%
  select(r.squared, sigma, nobs)
```

::: {.callout-tip icon="false"}
`sigma` = residual standard error = the typical size of a residual. Think of it as the average prediction error in units of $Y$.
:::

:::
:::::

## Interpreting $R^2$ in context

\

```{r}
#| echo: false
r2 <- round(model_glance$r.squared * 100, 1)
```

For our model: $R^2$ = `r r2`%

\

::: {.callout-important icon="false"}
### Interpretation

BMI explains approximately `r r2`% of the variability in systolic blood pressure in this NHANES sample. The remaining `r 100 - r2`% of variability is due to other factors not included in the model — age, sex, medications, genetics, sodium intake, stress, and so on.

An $R^2$ of `r r2`% sounds modest, but in a complex biological system with a single predictor, this is a meaningful association. **Context matters.**
:::

\

::: {.callout-warning icon="false"}
A high $R^2$ does not validate the model. It is possible to have a high $R^2$ with a curved relationship that a linear model fits poorly. **Always check the LINE conditions first.**
:::

# Part 4: Inference and Prediction

## What are we testing?

\

We have an estimated slope $\hat{\beta}_1$. But is the true population slope $\beta_1$ actually different from zero?

\

$$H_0: \beta_1 = 0 \qquad \text{vs.} \qquad H_A: \beta_1 \neq 0$$

\

::: {.callout-note icon="false"}
### Why $\beta_1 = 0$ matters

If $\beta_1 = 0$, the regression line is flat — $X$ tells us nothing about $Y$. Testing $H_0: \beta_1 = 0$ is equivalent to testing whether there is **any linear association** between the two variables.
:::

\

The test statistic is a $t$-statistic:

$$t = \frac{\hat{\beta}_1 - 0}{SE_{\hat{\beta}_1}} = \frac{\hat{\beta}_1}{SE_{\hat{\beta}_1}}$$

which follows a $t$-distribution with $df = n - 2$ under $H_0$.

## Inference for slope $\beta_1$: all 6 steps

**Step 1: Hypotheses**

$$H_0: \beta_1 = 0 \quad \text{(no linear association between BMI and systolic BP)}$$
$$H_A: \beta_1 \neq 0 \quad \text{(a linear association exists)}$$

**Step 2: Significance level** &nbsp; $\alpha = 0.05$

**Step 3: Check assumptions** &nbsp; LINE conditions — checked via diagnostic plots on previous slides ✓

**Step 4: Test statistic**

```{r}
b1_row <- tidy(model_bp) %>% filter(term == "BMI")
(t_stat <- b1_row$statistic)
```

$$t = \frac{\hat{\beta}_1}{SE_{\hat{\beta}_1}} = \frac{`r round(b1_row$estimate, 3)`}{`r round(b1_row$std.error, 3)`} = `r round(t_stat, 2)`$$

## Inference for slope $\beta_1$ (continued)

**Step 5: p-value**

```{r}
(p_val <- b1_row$p.value)
```

Or directly from the regression table:

```{r}
tidy(model_bp, conf.int = TRUE)
```

**Step 6: Conclusion**

We reject $H_0$ at the $\alpha = 0.05$ level (t = `r round(t_stat, 2)`, df = `r n_bp - 2`, p `r if(p_val < 0.001) "< 0.001" else paste("=", round(p_val, 3))`). There is strong evidence of a positive linear association between BMI and systolic blood pressure.

## Confidence interval for $\beta_1$

\

The 95% confidence interval for the population slope uses the same logic as all our previous CIs:

$$\hat{\beta}_1 \pm t^*_{n-2} \cdot SE_{\hat{\beta}_1}$$

where $t^*_{n-2}$ is the critical value from the $t$-distribution with $n - 2$ degrees of freedom.

\

```{r}
tidy(model_bp, conf.int = TRUE) %>%
  filter(term == "BMI") %>%
  select(term, estimate, conf.low, conf.high)
```

\

::: {.callout-important icon="false"}
### Interpretation

```{r}
#| echo: false
ci_lo <- round(tidy(model_bp, conf.int = TRUE) %>% filter(term == "BMI") %>% pull(conf.low), 3)
ci_hi <- round(tidy(model_bp, conf.int = TRUE) %>% filter(term == "BMI") %>% pull(conf.high), 3)
```

For every 1 kg/m² increase in BMI, systolic blood pressure is expected to increase by `r b1` mmHg, on average (95% CI: `r ci_lo` to `r ci_hi` mmHg).

Because the CI does not include 0, this is consistent with our hypothesis test conclusion: there is evidence of a positive linear association.
:::

## Prediction: using the model for new observations

\

Once we have a regression line, we can plug in any $X$ value to get a predicted $Y$:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 \cdot X^*$$

\

```{r}
new_data <- data.frame(BMI = c(25, 30, 35))

# Point predictions
predict(model_bp, newdata = new_data)
```

\

::: {.callout-tip icon="false"}
### What the confidence band in `geom_smooth` actually means

When you write `geom_smooth(method = "lm", se = TRUE)`, the shaded band is a **95% confidence interval for the mean response** at each value of $X$ — it answers: *"What is the average systolic BP for all people at this BMI?"*

The band is narrowest near $\bar{X}$ (the mean BMI) and widens toward the extremes. Predictions are less precise far from the center of the data.
:::

## Prediction intervals: a preview

\

There's a second type of interval you'll encounter — the **prediction interval** — which answers a different question:

\

| | Question | Accounts for |
|---|---|---|
| **CI for mean response** | What's the *average* BP for people with BMI = 30? | Uncertainty about the line |
| **Prediction interval** | What BP would *this specific patient* have? | Uncertainty about the line **+** individual variation |

\

Prediction intervals are always **wider** than confidence intervals for the mean, because individuals vary around the mean even if we knew the line perfectly.

\

::: {.callout-note icon="false"}
### Coming in BSTA 512

The mechanics of constructing and interpreting prediction intervals — along with multiple regression, model selection, and diagnostics for more complex models — are covered in depth in BSTA 512. You now have the conceptual foundation to hit the ground running there.
:::

# Wrap-Up

## The full SLR workflow

\

::::: columns
::: {.column width="55%"}

**1. Visualize**

```r
ggplot(data, aes(x = X, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm")
```

**2. Fit**

```r
model <- lm(Y ~ X, data = data)
```

**3. Coefficients + inference**

```r
tidy(model, conf.int = TRUE)
```

**4. Model fit**

```r
glance(model)   # R², sigma
```

:::

::: {.column width="5%"}
:::

::: {.column width="40%"}

**5. Diagnostics**

```r
aug <- augment(model)

# Residual plot
ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0)

# QQ plot
ggplot(aug, aes(sample = .resid)) +
  stat_qq() + stat_qq_line()
```

**6. Predict**

```r
predict(model,
  newdata = new_data)
```

:::
:::::

## Where this fits in the bigger picture

\

::::: columns
::: {.column width="55%"}

**What we've learned this semester:**

- Describing data (distributions, summaries, visualization)
- Inference for one group (means, proportions)
- Inference for two groups (t-tests, difference in proportions)
- Inference for 3+ groups (ANOVA, Kruskal-Wallis)
- Inference for categorical associations (chi-square, Fisher's)
- Inference for linear associations **(today)**

**Where this leads:**

- Multiple regression (add more predictors)
- Logistic regression (binary outcome)
- Survival analysis, mixed models, causal inference

These are the tools of modern biomedical research. BSTA 512 takes it from here.

:::

::: {.column width="5%"}
:::

::: {.column width="40%"}

::: {.callout-tip icon="false"}
### The unified thread

Every method we've learned is asking the same fundamental question:

**What does the data tell us about the population, and how confident are we?**

The tools change. The logic — estimate, quantify uncertainty, test — stays the same.
:::

:::
:::::

\

**Next: Lesson 19 — Course Review and Final Exam Preparation**
