---
title: "R Module 4: Tables and Formatted Output"
subtitle: "BMSC 620 | R Workflow Series"
author: "Emile Latour"
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v6.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | Tables and Formatted Output"
    html-math-method: mathjax
    chalkboard: true
    progress: true
    code-line-numbers: true
    code-copy: true
    highlight-style: github
    fig-align: center
    incremental: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

knitr::opts_chunk$set(dpi = 300)

library(tidyverse)
library(oibiostat)
library(knitr)
library(gt)
library(gtsummary)

data("nhanes.samp")

nhanes.samp <- nhanes.samp %>%
  mutate(
    Gender    = factor(Gender, levels = c("female", "male"),
                       labels = c("Female", "Male")),
    HealthGen = factor(as.character(HealthGen),
                       levels = c("Excellent", "Vgood", "Good", "Fair", "Poor"),
                       labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))
  )
```

## What We're Learning Today {.section-slide}

::: {.learning-objectives}
By the end of this module, you will be able to:

1. Use `knitr::kable()` to produce clean simple tables
2. Use the `gt` package to build **publication-quality** formatted tables
3. Apply column labels, number formatting, and captions inside tables
4. Use `gtsummary::tbl_summary()` to generate a **Table 1** with minimal code
:::

::: {.callout-note}
These skills close the gap between raw R output and results you can actually put in a report.
:::

---

## The Problem We're Solving

This is what raw R output looks like:

```{r}
#| echo: true

nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = mean(BMI, na.rm = TRUE),
    sd   = sd(BMI, na.rm = TRUE)
  )
```

---

## The Problem We're Solving

It's correct. But it is **not a result**:

\

- Column names are code names, not readable labels
- Numbers have too many decimal places
- There is no title or caption
- It would look terrible in a report

. . .

\

::: {.callout-important icon="false"}
### The principle for this module

A wall of raw R output is not a result. A table is. The gap between `print(df)` and a manuscript-ready table is smaller than you think.
:::

---

## Part 1: `knitr::kable()` {.section-slide}

---

## The Simplest Upgrade

`kable()` from the `knitr` package converts any data frame into a formatted
HTML or PDF table in one line:

```{r}
nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = round(mean(BMI, na.rm = TRUE), 1),
    sd   = round(sd(BMI, na.rm = TRUE), 1)
  ) %>%
  knitr::kable()
```

---

## `kable()` with Options


::: columns
::: {.column width="60%"}
```{r}
nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = round(mean(BMI, na.rm = TRUE), 1),
    sd   = round(sd(BMI, na.rm = TRUE), 1)
  ) %>%
  knitr::kable(
    col.names = c("Sex", "N", "Mean BMI", "SD"),
    caption   = "BMI by sex in the NHANES sample",
    align     = c("l", "c", "c", "c")
  )
```
:::
::: {.column width="40%"}

\

\

\

\

\

\

\

\

\

\

| Argument | Purpose |
|---|---|
| `col.names` | Override column names |
| `caption` | Table caption |
| `align` | Column alignment: `"l"`, `"c"`, `"r"` |
| `digits` | Decimal places (applies globally) |
:::
:::




---

## When to Use `kable()`

`kable()` is the right tool when you need a quick, clean table:

- Simple summary statistics
- Frequency tables from `tabyl()`
- Model coefficient tables from `tidy()`

\

For **publication-quality tables** with custom formatting, colors, or merged
cells — use `gt`.

---

## Part 2: The `gt` Package {.section-slide}

---

## Why `gt`?

`gt` ("grammar of tables") gives you fine-grained control over every element
of a table — similar to how `ggplot2` gives you control over every element of a plot.

\

::: {.columns}
::: {.column width="50%"}
**`kable()`**

- One function
- Works for quick clean tables
- Limited formatting options
:::

::: {.column width="50%"}
**`gt`**

- Layered grammar (like ggplot2)
- Column labels, formatting, spanners, footnotes
- Suitable for manuscripts and reports
:::
:::

\

::: {.callout-tip icon="false"}
`gt` tables render beautifully in HTML documents. For PDF output, additional setup is needed — HTML is the target format for this course.
:::

---

## A Minimal `gt` Table

Start with `gt()` and add layers:

::: {.gt_table}
```{r}
nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = round(mean(BMI, na.rm = TRUE), 1),
    sd   = round(sd(BMI, na.rm = TRUE), 1)
  ) %>%
  gt()
```
:::

---

## Adding a Title and Column Labels


```{r}
#| eval: false
nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = round(mean(BMI, na.rm = TRUE), 1),
    sd   = round(sd(BMI, na.rm = TRUE), 1)
  ) %>%
  gt() %>%
  tab_header(
    title    = "BMI by Sex",
    subtitle = "NHANES sample (n = 200)"
  ) %>%
  cols_label(
    Gender = "Sex",
    n      = "N",
    mean   = "Mean BMI",
    sd     = "SD"
  )
```


## RESULT: Adding a Title and Column Labels 

\

::: {.gt_table}
```{r}
#| echo: false
nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = round(mean(BMI, na.rm = TRUE), 1),
    sd   = round(sd(BMI, na.rm = TRUE), 1)
  ) %>%
  gt() %>%
  tab_header(
    title    = "BMI by Sex",
    subtitle = "NHANES sample (n = 200)"
  ) %>%
  cols_label(
    Gender = "Sex",
    n      = "N",
    mean   = "Mean BMI",
    sd     = "SD"
  )
```
:::

---

## Formatting Numbers Inside the Table

`fmt_number()` formats specific columns — no more `round()` before the table. Now you keep raw values in the data and format only at the display stage.

```{r}
#| eval: false

nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = mean(BMI, na.rm = TRUE),
    sd   = sd(BMI, na.rm = TRUE)
  ) %>%
  gt() %>%
  tab_header(title = "BMI by Sex", subtitle = "NHANES sample (n = 200)") %>%
  cols_label(Gender = "Sex", n = "N", mean = "Mean BMI", sd = "SD") %>%
  fmt_number(columns = c(mean, sd), decimals = 1)
```

::: {.gt_table}
```{r}
#| echo: false

nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = n(),
    mean = mean(BMI, na.rm = TRUE),
    sd   = sd(BMI, na.rm = TRUE)
  ) %>%
  gt() %>%
  tab_header(title = "BMI by Sex", subtitle = "NHANES sample (n = 200)") %>%
  cols_label(Gender = "Sex", n = "N", mean = "Mean BMI", sd = "SD") %>%
  fmt_number(columns = c(mean, sd), decimals = 1)
```
:::

---

## Key `gt` Formatting Functions

| Function | Purpose |
|---|---|
| `tab_header()` | Title and subtitle |
| `cols_label()` | Human-readable column names |
| `fmt_number()` | Format numeric columns (decimals, comma separators) |
| `fmt_percent()` | Format as percentages |
| `tab_footnote()` | Add footnotes to cells or columns |
| `tab_spanner()` | Group columns under a shared label |
| `cols_align()` | Column alignment |
| `tab_style()` | Bold, color, or style specific cells |

---

## A Multi-Variable Summary Table

```{r}
#| eval: false


nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n       = n(),
    bmi_m   = mean(BMI, na.rm = TRUE),
    bmi_sd  = sd(BMI, na.rm = TRUE),
    sbp_m   = mean(BPSysAve, na.rm = TRUE),
    sbp_sd  = sd(BPSysAve, na.rm = TRUE)
  ) %>%
  gt() %>%
  tab_header(
    title    = "BMI and Systolic Blood Pressure by Sex",
    subtitle = "NHANES sample (n = 200)"
  ) %>%
  cols_label(
    Gender = "Sex",    n = "N",
    bmi_m  = "Mean",   bmi_sd = "SD",
    sbp_m  = "Mean",   sbp_sd = "SD"
  ) %>%
  fmt_number(columns = c(bmi_m, bmi_sd, sbp_m, sbp_sd), decimals = 1) %>%
  tab_spanner(label = "BMI (kg/m²)",         columns = c(bmi_m, bmi_sd)) %>%
  tab_spanner(label = "Systolic BP (mmHg)",  columns = c(sbp_m, sbp_sd))
```

## A Multi-Variable Summary Table

::: {.gt_table}
```{r}
#| echo: false

nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n       = n(),
    bmi_m   = mean(BMI, na.rm = TRUE),
    bmi_sd  = sd(BMI, na.rm = TRUE),
    sbp_m   = mean(BPSysAve, na.rm = TRUE),
    sbp_sd  = sd(BPSysAve, na.rm = TRUE)
  ) %>%
  gt() %>%
  tab_header(
    title    = "BMI and Systolic Blood Pressure by Sex",
    subtitle = "NHANES sample (n = 200)"
  ) %>%
  cols_label(
    Gender = "Sex",    n = "N",
    bmi_m  = "Mean",   bmi_sd = "SD",
    sbp_m  = "Mean",   sbp_sd = "SD"
  ) %>%
  fmt_number(columns = c(bmi_m, bmi_sd, sbp_m, sbp_sd), decimals = 1) %>%
  tab_spanner(label = "BMI (kg/m²)",         columns = c(bmi_m, bmi_sd)) %>%
  tab_spanner(label = "Systolic BP (mmHg)",  columns = c(sbp_m, sbp_sd))
```
:::

---

## Part 3: Table 1 with `gtsummary` {.section-slide}

---

## What Is a Table 1?

In clinical research, **Table 1** is the first table in a manuscript — it
describes the sample by group:

\

::: {.callout-note icon="false"}
### Every clinical paper has one

"Table 1: Baseline characteristics of study participants by treatment group"

For continuous variables: mean (SD) or median (IQR)

For categorical variables: n (%)
:::

\

Writing Table 1 by hand is tedious and error-prone. `gtsummary` generates it automatically.

---

## `tbl_summary()` — Your First Table 1

::: {.gt_table}
```{r}
nhanes.samp %>%
  select(Gender, Age, BMI, BPSysAve) %>%
  tbl_summary()
```
:::

That's it. `tbl_summary()` detects variable types automatically:

- Continuous → median (IQR) by default
- Categorical → n (%)

---

## Stratifying by a Group Variable

Add `by = Gender` to split the table into columns:

```{r}
nhanes.samp %>%
  select(Gender, Age, BMI, BPSysAve) %>%
  tbl_summary(by = Gender)
```

---

## Customizing `tbl_summary()`

```{r}
#| output-location: slide

nhanes.samp %>%
  select(Gender, Age, BMI, BPSysAve, HealthGen) %>%
  tbl_summary(
    by        = Gender,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits    = all_continuous() ~ 1,
    label     = list(
      Age       ~ "Age (years)",
      BMI       ~ "BMI (kg/m²)",
      BPSysAve  ~ "Systolic BP (mmHg)",
      HealthGen ~ "Self-Reported Health"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```

---

## `tbl_summary()` Key Arguments

| Argument | Purpose |
|---|---|
| `by` | Stratify columns by a group variable |
| `statistic` | Control the summary format per variable type |
| `digits` | Decimal places for continuous variables |
| `label` | Human-readable variable labels |
| `include` | Which variables to include |

\

**Post-processing functions:**

| Function | Purpose |
|---|---|
| `add_overall()` | Add an "Overall" column |
| `add_p()` | Add p-values comparing groups |
| `bold_labels()` | Bold the row variable labels |

---

## Putting It All Together {.section-slide}

---

## Choosing the Right Tool

\

| Situation | Tool |
|---|---|
| Quick check of output during analysis | `print()` / console |
| Simple table in a homework or report | `knitr::kable()` |
| Formatted multi-variable table for a report | `gt` |
| Table 1 with group comparisons | `gtsummary::tbl_summary()` |
| Model output (coefficients, CIs) | `gtsummary::tbl_regression()` |

\

::: {.callout-tip icon="false"}
You don't need to memorize every function — you need to know which tool fits the situation, and know enough to look up the rest.
:::

---

## Key Takeaways

::: {.columns}
::: {.column width="60%"}
**`kable()`**

- `knitr::kable()` — one function, instantly clean
- `col.names`, `caption`, `align`, `digits`

**`gt`**

- `gt()` → `tab_header()` → `cols_label()` → `fmt_number()`
- `tab_spanner()` for grouped column headers
- Keep raw numbers in data; format only at display stage

**`gtsummary`**

- `tbl_summary(by = ...)` generates Table 1 automatically
- `statistic` controls mean (SD) vs median (IQR)
- `label` gives clean variable names
- `add_overall()` and `add_p()` for the finishing touches
:::

::: {.column width="40%"}
::: {.callout-tip icon="false"}
**The rule:**

Raw R output is for you. A table is for your reader. Every number that appears in a report should be in a table or inline R — never copy-pasted.
:::
:::
:::

---

## References

\

**`knitr::kable()`**

- [knitr documentation: kable](https://bookdown.org/yihui/rmarkdown-cookbook/kable.html)

\

**`gt`**

- [gt package documentation](https://gt.rstudio.com/)
- [gt cookbook](https://themockup.blog/static/resources/gt-cookbook.html) — Thomas Mock

\

**`gtsummary`**

- [gtsummary documentation](https://www.danieldsjoberg.com/gtsummary/)
- [tbl_summary() tutorial](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html)
