---
title: "BMSC 620 — Homework 0"
author: "Your Name"
format: html
execute:
  echo: true
---

::: {.callout-note}
**File naming and submission**

Please use the following file naming convention:

`Lastname_FirstInitial_HW##`

Examples:

- Smith_J_HW01.qmd
- Smith_J_HW01.html


When submitting Homework 0, upload **both**:

- the `.qmd` file
- the rendered `.html` file
:::

## Homework 0

This assignment is intended to help you get comfortable working in RStudio
and using Quarto documents.

This assignment is graded for completion and effort, not correctness.

---

## Part 1: Markdown practice

Edit this section to demonstrate the following:

- Bold text
- Italic text
- A bulleted list
- A numbered list

You may replace the text below with your own examples.

- Example bullet
- Another bullet

1. Example item
2. Another item

---

## Part 2: R code chunks

Below is an example R code chunk.

```{r}
# This is a comment

2 + 2

```

### Tasks

1.	Modify the code chunk above so that it:
  - Includes at least one additional comment
  - Runs without errors
2. Add a second R code chunk below that:
  - Creates a numeric vector with at least 5 numbers 
  - Computes the mean of that vector using mean()
  - I've included an example below as one possible solution for reference. The submitted code should be your own.

### Example (for reference only)

```{r}

# Example

# Create a numeric vector
x <- c(3, 7, 10, 4, 6)

# Compute the mean
mean(x)

```

```{r}

# Your code here

```

## Part 3: Final check

Before you are finished, confirm that:

- The document renders to HTML without errors
- All R code chunks run successfully
- Your name appears at the top of the document

