---
title: "BMSC 620 — Homework 1"
author: "Your Name"
format: 
  html:
    embed-resources: true
execute:
  echo: true
---

::: {.callout-note}
**File naming and submission**

Please use the following file naming convention:

`Lastname_FirstInitial_HW##`

Examples:

- Smith_J_HW01.qmd
- Smith_J_HW01.html


When submitting Homework 1, upload **both**:

- the `.qmd` file
- the rendered `.html` file
:::

## Homework 1

This assignment reviews:

- writing and running code in Quarto
- summarizing numeric and categorical variables
- frequency tables and contingency tables
- bar plots, histograms, and box plots

Unless stated otherwise, show your work using R code chunks.

---


## Part 0: Setup

1. In the chunk below, load the packages you need for this homework.
2. Run the chunk to confirm there are no errors.

```{r}

# NOTE: If you don't already have the oibiostat package installed, then uncomment and run the two lines below. Then re-comment them and check installation with `library(oibiostat)`.
# install.packages("devtools")
# devtools::install_github("OI-Biostat/oi_biostat_data", force = TRUE)

library(oibiostat)

```

## Part 1: R objects and vectors

### 1A. Calculator practice

Create code that computes each of the following (one line each):

- 12 / 5
- 3^4
- sqrt(81)
- 4^3 - 2 * 7 + 9 / 2

```{r}
# Your code here


```

### 1B. Create vectors and summarize

1. Create a vector named x containing the numbers: 4, 9, 2, 10, 7, 7
2. Compute the mean and standard deviation of x
3. Compute the median and IQR of x

```{r}

```


### 1C. Missing values

1. Create a vector named y containing: 2, 5, NA, 7, 9
2. Compute mean(y). What happens?
3. Compute the mean again ignoring missing values

Write 1 to 2 sentences explaining what na.rm = TRUE does.

```{r}

```

**Your explanation:**

## Part 2: Inspecting a dataset (nhanes.samp)

In this homework, you will work with the dataset **nhanes.samp**, which is included in the `oibiostat` package.

The National Health and Nutrition Examination Survey (NHANES) is a large, ongoing study conducted in the United States to assess the health and nutritional status of adults and children. The dataset used here is a **sample of 200 individuals** from NHANES, provided for instructional purposes.

To load the dataset, run:

```{r}
library(oibiostat)
data("nhanes.samp")

```

### 2A. Basic inspection

Run the following functions:

- dim(nhanes.samp)
- names(nhanes.samp)
- str(nhanes.samp)
- head(nhanes.samp)

```{r}

```


Answer the questions below in plain text.

1. How many rows and columns does nhanes.samp have?
2. Name two categorical variables? (i.e. `Factor` variable type)
3. Give one example of a numeric variable. (`int` or `num`)

Answers:
1.
2.
3.

### 2B. Accessing a single variable with $

1. Use $ to pull out `Age`
2. Compute the mean, sd, median, and IQR of `Age`

```{r}

```

### 2C. Working with `NA` values. 

We can see that some variables in the dataset have missing values (`NA`) like the variable `Height`.

```{r}
summary(nhanes.samp$Height)
```

1. Use $ to pull out `Height`
2. Compute the mean, sd, median, and IQR of `Height`. Hint: you will need to remove the `NA` values by adding the argument `na.rm = TRUE` to the function calls.

```{r}

```



## Part 3: Summarizing one categorical variable

We will summarize `Gender`

### 3A. Counts and proportions

1. Use table() to get counts of `Gender`
2. Use prop.table() to get proportions of `Gender`
3. Convert proportions to percentages

Hint: `table(nhanes.samp$Gender)` and `prop.table(...)`

```{r}

```


### 3B. Bar plot

Make a bar plot of `Gender` using the `barplot` function.

```{r}

```

## Part 4: Two categorical variables (contingency tables)

We can use the `table()` function to examine the relationships between two categorical variables. Here is an example using `Gender` and `Race1`:

```{r}
table(nhanes.samp$Gender, nhanes.samp$Race1)   # counts
prop.table(table(nhanes.samp$Gender, nhanes.samp$Race1), margin = 1)  # row proportions
prop.table(table(nhanes.samp$Gender, nhanes.samp$Race1), margin = 2)  # column proportions
```

Using **row percentages (margin = 1)**, where each row sums to 100% and Gender is held fixed, we see that among females in the data set, 7 (6.8%) are Black, 8 (7.8%) Hispanic, 9 (8.8%) Mexican, 65 (63.7%) White, and 13 (12.7%) Other.

Using **column percentages (margin = 2)**, where each column sums to 100% and Race is held fixed, we see that among the largest racial group, White, 53.3% are female and 46.7% are male.

### 4A. Contingency table (counts)

Choose either the `Diabetes` or `SmokeNow` variable in the data set. Create a contingency table similar to above with `Gender` and the variable that you choose.

```{r}

```

### 4B. Row proportions vs column proportions

1. Create a row proportion table (each row sums to 1)
2. Create a column proportion table (each column sums to 1)

Hint: 

- `prop.table(table(nhanes.samp$Gender, YOUR VARIABLE), margin = 1)`
- `prop.table(table(nhanes.samp$Gender, YOUR VARIABLE), margin = 2)`

```{r}

```


### 4C. Interpretation

Answer each question in 1 to 2 sentences.

1. Provide 1 or 2 short sentences describing the *row* percentages.
2. Provide 1 or 2 short sentences describing the *column* percentages.

Answers:
1.
2.


## Part 5: Visual summaries for numeric variables

### 5A. Histogram

Make a histogram of `BMI`.

Using the `hist()` function, you can add:

- a title by adding an argument `main = "Histogram of BMI"`
- and a label for the x-axis using `xlab = "BMI"`
- When in doubt about arguments to functions, use the `?` then the function name to get help. i.e. type in the console `?hist` to get help for the `hist()` function.

If you get an error due to missing values, use `na.omit()` or remove `NA` values before calling `hist()`.

```{r}

```

### 5B. Box plot (overall)

Make a single box plot of `Height` using `boxplot()`. Include a title "Box plot of Height" and a label for the y-axis (`ylab = "Height"`). Get help using `?boxplot` if you need it.

NOTE: We saw earlier that `Height` contained `NA` values. `boxplot()` will ignore `NA` values automatically.

```{r}

```

### 5C. Box plot by group

Make a box plot of `Height` by `Gender`.

If you look at the first example in the help for the boxplot function `?boxplot`, they give an example of a box plot of `count` by `spray` using the `InsectSprays` dataset.

```{r}
#| eval: false
boxplot(count ~ spray, data = InsectSprays, col = "lightgray")
```

Follow their example to make your `Height` by `Gender` box plot with the `nhanes.samp` data.

```{r}

```

Which group has a higher median height? How can you tell from the figure? Are there any outliers?

Answer:

## Final check

Before submitting, confirm that:

- The document renders to HTML without errors
- All code chunks run successfully
- Your name appears at the top
- You uploaded both the .qmd and .html files

