---
title: "R Module 3: Visual Communication with ggplot2"
subtitle: "BMSC 620 | R Workflow Series"
author: "Emile Latour"
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v5.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | Visual Communication with ggplot2"
    html-math-method: mathjax
    chalkboard: true
    progress: true
    code-line-numbers: true
    code-copy: true
    highlight-style: github
    fig-align: center
    incremental: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

knitr::opts_chunk$set(dpi = 300)

library(tidyverse)
library(oibiostat)
library(scales)
library(patchwork)

data("nhanes.samp")

# Set up HealthGen as a properly ordered factor with clean labels
nhanes.samp <- nhanes.samp %>%
  mutate(
    HealthGen = factor(as.character(HealthGen),
                       levels = c("Excellent", "Vgood", "Good", "Fair", "Poor"),
                       labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))
  )
```

## What We're Learning Today {.section-slide}

::: {.learning-objectives}
By the end of this module, you will be able to:

1. Apply and customize **themes** to control the overall look of a plot
2. Use **colorblind-accessible palettes** with `scale_fill/color_manual()`
3. **Format axis labels** with `scale_x/y_continuous()` and `scales` functions
4. Write informative **plot labels** using `labs()` well
5. Add **contextual annotations** directly to plots with `annotate()`
:::

::: {.callout-note}
These skills apply every time you put a plot in a report, a poster, or a manuscript.
:::



## The Problem We're Solving

Here is what ggplot2 gives you by default:

```{r}
#| echo: true
#| fig-width: 8
#| fig-height: 4.5

ggplot(nhanes.samp, aes(x = HealthGen, y = BPSysAve)) +
  geom_boxplot()
```



## The Problem We're Solving

\

It runs. It's technically correct. But it **fails as communication**:

\

- The grey background competes with the data
- All boxes are the same color — no visual encoding of groups
- Axis labels say `HealthGen` and `BPSysAve` — meaningless to a reader
- No title. No context. No reference lines.

. . .

\

::: {.callout-important icon="false"}
### The principle for this module

A plot that requires the reader to mentally decode it has failed. Our job is to make the finding obvious.
:::



# Part 1: Themes {.section-slide}



## Built-In Themes

ggplot2 ships with several themes. The default is `theme_grey()`.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

base_plot <- nhanes.samp %>% filter(!is.na(BMI)) %>% 
  ggplot(aes(x = BMI)) +
  geom_histogram(bins = 25, fill = "#0072B2", color = "white") +
  labs(x = "BMI (kg/m²)", y = "Count")

p1 <- base_plot + theme_grey()   + labs(subtitle = "theme_grey()   (default)")
p2 <- base_plot + theme_bw()     + labs(subtitle = "theme_bw()")
p3 <- base_plot + theme_minimal()+ labs(subtitle = "theme_minimal()")
p4 <- base_plot + theme_classic()+ labs(subtitle = "theme_classic()")

p1 + p2 + p3 + p4
```

For biomedical reporting, `theme_bw()` and `theme_classic()` are the most common choices.



## Applying a Theme

Add the theme as a layer — it goes at the end:

```{r}
#| fig-width: 7
#| fig-height: 4
nhanes.samp %>% filter(!is.na(BMI)) %>% 
  ggplot(aes(x = BMI)) +
  geom_histogram(bins = 25, fill = "#0072B2", color = "white") +
  labs(x = "BMI (kg/m²)", y = "Count",
       title = "Distribution of BMI") +
  theme_bw(base_size = 14)
```

`base_size` scales all text elements proportionally. The default is 11 — increase it for presentations and posters.



## Fine-Tuning with `theme()`

`theme()` lets you override individual elements after setting a base theme:

```{r}
#| eval: false

ggplot(...) +
  geom_histogram(...) +
  theme_bw(base_size = 14) +
  theme(
    plot.title    = element_text(face = "bold", size = 16),
    axis.title    = element_text(size = 13),
    panel.grid.minor = element_blank(),      # remove minor gridlines
    legend.position  = "bottom"              # move legend
  )
```

\

::: {.callout-tip icon="false"}
### The pattern

`theme_bw()` sets the foundation. `theme()` adjusts the details. Always put `theme()` **after** your base theme — it overrides rather than replaces.
:::



## Useful `theme()` Arguments

\

| Argument | What it controls |
|---|---|
| `plot.title` | Title text style |
| `plot.subtitle` | Subtitle text style |
| `axis.title` | Both axis labels |
| `axis.text` | Tick mark labels |
| `legend.position` | `"right"`, `"bottom"`, `"top"`, `"none"` |
| `panel.grid.minor` | Minor gridlines |
| `panel.grid.major` | Major gridlines |
| `strip.text` | Facet panel labels |

\

Use `element_text(face = "bold", size = 14, color = "grey20")` to style text elements.

Use `element_blank()` to remove an element entirely.



# Part 2: Colors and Palettes {.section-slide}



## The Default Color Problem

```{r}
#| fig-width: 7
#| fig-height: 4.5
#| echo: true
nhanes.samp %>% filter(!is.na(HealthGen), !is.na(BPSysAve)) %>% 
  ggplot(aes(x = HealthGen, y = BPSysAve, fill = HealthGen)) +
  geom_boxplot() +
  theme_bw(base_size = 14)
```

The default ggplot2 palette is **not colorblind-accessible** — approximately 8% of males have some form of color vision deficiency.



## The Okabe-Ito Palette

Designed to be distinguishable under the most common forms of color vision deficiency:

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 2.5

oi_colors <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2",
               "#D55E00", "#CC79A7", "#F0E442", "#000000")
oi_names  <- c("Orange", "Sky Blue", "Green", "Blue",
               "Vermillion", "Pink", "Yellow", "Black")

tibble(color = oi_colors, name = oi_names, x = seq_along(oi_colors)) %>%
  ggplot(aes(x = x, y = 1, fill = color)) +
  geom_tile(width = 0.9, height = 0.7) +
  geom_text(aes(label = name), y = 0.55, size = 4.5, color = "white",
            fontface = "bold") +
  geom_text(aes(label = color), y = 1.45, size = 3.5, color = "grey30") +
  scale_fill_identity() +
  scale_x_continuous(expand = c(0.02, 0)) +
  theme_void() +
  labs(title = "The Okabe-Ito Palette (8 colors)") +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
```

Use `scale_fill_manual()` or `scale_color_manual()` to apply it.

\

::: {.callout-tip icon="false"}
The package [**ggokabeito**](https://malcolmbarrett.github.io/ggokabeito/) wraps 
the Okabe-Ito palette into `scale_fill_okabe_ito()` and `scale_color_okabe_ito()` 
— no hex codes needed. 
:::



## `scale_fill_manual()` — For Filled Geoms

Use this for `geom_bar()`, `geom_boxplot()`, `geom_histogram()`, `geom_area()`:

```{r}
#| fig-width: 8
#| fig-height: 4.5

nhanes.samp %>% filter(!is.na(HealthGen), !is.na(BPSysAve)) %>% 
  ggplot(aes(x = HealthGen, y = BPSysAve, fill = HealthGen)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73",
                                "#0072B2", "#D55E00"),
                    guide = "none") +
  theme_bw(base_size = 14)
```

`guide = "none"` suppresses the legend when the x-axis already encodes the grouping.



## `scale_color_manual()` — For Outlined/Point Geoms

Use this for `geom_point()`, `geom_line()`, `geom_boxplot(color = ...)`:

```{r}
#| fig-width: 8.5
#| fig-height: 4.5

nhanes.samp %>% filter(!is.na(Gender), !is.na(BMI), !is.na(BPSysAve)) %>% 
  ggplot(aes(x = BMI, y = BPSysAve, color = Gender)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("#E69F00", "#0072B2")) +
  theme_bw(base_size = 14)
```



::: {.callout-tip icon="false"}
The number of colors in `values` must match the number of groups. Supply them in the same order as the factor levels.
:::



# Part 3: Axes and Scales {.section-slide}



## `scale_x/y_continuous()`

Use `scale_x_continuous()` or `scale_y_continuous()` to control continuous axes:

```{r}
#| eval: false

scale_y_continuous(
  limits = c(70, 200),                           # axis range
  breaks = seq(80, 200, by = 20),                # where tick marks appear
  labels = scales::number_format(accuracy = 1)   # how labels are formatted
)
```

\

| Argument | Purpose |
|---|---|
| `limits` | Set the min and max shown |
| `breaks` | Tick mark positions |
| `labels` | How to format the tick labels |
| `expand` | Padding at the edges |



## Formatting Axis Labels with `scales`

The `scales` package gives you format functions for labels:

```{r}
#| fig-width: 8
#| fig-height: 4
nhanes.samp %>% filter(!is.na(BMI)) %>% 
  ggplot(aes(x = BMI)) +
  geom_histogram(bins = 25, fill = "#0072B2", color = "white") +
  scale_x_continuous(
    breaks = seq(15, 70, by = 5),
    labels = scales::number_format(accuracy = 1)
  ) +
  theme_bw(base_size = 14)
```



## Formatting as Percentages

`scales::percent_format()` is useful when your axis represents a proportion:

```{r}
#| fig-width: 8
#| fig-height: 4

nhanes.samp %>%
  filter(!is.na(HealthGen)) %>%
  count(HealthGen) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = HealthGen, y = prop)) +
  geom_col(fill = "#0072B2") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_bw(base_size = 14)
```


# Part 4: Labels Done Well 


## `labs()` — Your Communication Toolkit

Every element of `labs()` serves a purpose:

```{r}
#| eval: false

labs(
  title    = "The main finding, stated clearly",
  subtitle = "Sample description or key context",
  x        = "Axis label with units",
  y        = "Axis label with units",
  color    = "Legend title for color aesthetic",
  fill     = "Legend title for fill aesthetic",
  caption  = "Data source or methodological note"
)
```

\

::: {.callout-tip icon="false"}
### What makes a good title?

State the finding, not the variable name. **"BMI Distribution"** is a label. **"BMI is right-skewed, with most adults between 20 and 35 kg/m²"** is a finding.

For exploratory plots (in homework), descriptive titles are fine. For manuscripts and presentations, state the finding.
:::



## Labels: Before and After

::: {.columns}
::: {.column width="50%"}
**Minimal labels:**

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5

nhanes.samp %>% filter(!is.na(Gender), !is.na(BPSysAve)) %>% 
  ggplot(aes(x = Gender, y = BPSysAve, fill = Gender)) +
  geom_boxplot(alpha = 0.8, 
               width = 0.4) +
  scale_fill_manual(values = c("#E69F00", "#0072B2"), guide = "none") +
  theme_bw(base_size = 14)
```
:::

::: {.column width="50%"}
**Informative labels:**

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5

nhanes.samp %>% filter(!is.na(Gender), !is.na(BPSysAve)) %>% 
  mutate(Gender = factor(Gender, 
                         levels = c("female", "male"), 
                         labels = c("Female", "Male"))) %>% 
  ggplot(aes(x = Gender, y = BPSysAve, fill = Gender)) +
  geom_boxplot(alpha = 0.8, 
               width = 0.4) +
  scale_fill_manual(values = c("#E69F00", "#0072B2"), guide = "none") +
  labs(
    title    = "Systolic blood pressure is higher in males",
    subtitle = "NHANES sample adults (n = 200)",
    x        = "Sex",
    y        = "Systolic Blood Pressure (mmHg)",
    caption  = "Data: NHANES sample from the oibiostat package"
  ) +
  theme_bw(base_size = 14)
```
:::
:::

Same data, same geom. The labels do the communication.



# Part 5: Annotate {.section-slide}



## Adding Context Directly to Plots

\

Sometimes the most important piece of information isn't in the data — it's a threshold, a benchmark, or a label that anchors the reader:

\

- A hypertension threshold line at SBP = 130 mmHg
- A reference line at BMI = 30 (obesity threshold)  
- A label pointing to an outlier or a notable group

\

Two tools:

- `geom_hline()` / `geom_vline()` — draw the reference line
- `annotate()` — add text, arrows, or shaded regions directly on the plot



## `geom_hline()` — Horizontal Reference Lines

```{r}
#| fig-width: 8
#| fig-height: 4.5

nhanes.samp %>% filter(!is.na(HealthGen), !is.na(BPSysAve)) %>% 
  ggplot(aes(x = HealthGen, y = BPSysAve, fill = HealthGen)) +
  geom_boxplot(alpha = 0.8) +
  geom_hline(yintercept = 130, linetype = "dashed",
             color = "grey40", linewidth = 0.8) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73",
                               "#0072B2", "#D55E00"),
                    guide = "none") +
  theme_bw(base_size = 14)
```




## `annotate("text")` — Text on the Plot

Add a text label at a specific data coordinate:

```{r}
#| fig-width: 8
#| fig-height: 4.5

nhanes.samp %>% filter(!is.na(HealthGen), !is.na(BPSysAve)) %>% 
ggplot(aes(x = HealthGen, y = BPSysAve, fill = HealthGen)) +
  geom_boxplot(alpha = 0.8) +
  geom_hline(yintercept = 130, linetype = "dashed",
             color = "grey40", linewidth = 0.8) +
  annotate("text", x = 1, y = 134,
           label = "Hypertension threshold (130 mmHg)",
           hjust = 0, size = 4, color = "grey30") +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73",
                                "#0072B2", "#D55E00"),
                    guide = "none") +
  theme_bw(base_size = 14)
```



## `annotate()` Arguments

```{r}
#| eval: false

annotate(
  geom  = "text",      # "text", "rect", "segment", "pointrange"
  x     = 1,           # x position (data coordinates)
  y     = 134,         # y position (data coordinates)
  label = "...",       # the text to display
  hjust = 0,           # horizontal justification: 0 = left, 1 = right
  vjust = 0,           # vertical justification: 0 = bottom, 1 = top
  size  = 4,           # text size
  color = "grey30"     # text color
)
```

\

::: {.callout-important icon="false"}
Positions in `annotate()` use **data coordinates**, not pixels or proportions. Set `x` and `y` to values that make sense on your axes.
:::



# Putting It All Together {.section-slide}



## Before: The Default Plot

```{r}
#| fig-width: 8
#| fig-height: 5

ggplot(nhanes.samp, aes(x = HealthGen, y = BPSysAve)) +
  geom_boxplot()
```

This is our starting point. Grey background, no colors, variable names as axis labels, no context.



## After: The Polished Plot

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 5

nhanes.samp %>% filter(!is.na(HealthGen), !is.na(BPSysAve)) %>% 
ggplot(aes(x = HealthGen, y = BPSysAve, fill = HealthGen)) +
  geom_boxplot(alpha = 0.85, outlier.shape = 21,
               outlier.size = 1.5, outlier.alpha = 0.5) +
  geom_hline(yintercept = 130, linetype = "dashed",
             color = "grey40", linewidth = 0.8) +
  annotate("text", x = 1, y = 134,
           label = "Hypertension threshold (130 mmHg)",
           hjust = 0, size = 4.5, color = "grey30") +
  scale_fill_manual(
    values = c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00"),
    guide  = "none"
  ) +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 1),
    limits = c(75, 180),
    breaks = seq(80, 180, by = 20)
  ) +
  labs(
    title    = "Systolic blood pressure increases with poorer self-reported health",
    subtitle = "NHANES sample adults (n = 200)",
    x        = "Self-Reported Health Status",
    y        = "Systolic Blood Pressure (mmHg)",
    caption  = "Data: NHANES sample from the oibiostat package. Dashed line = hypertension threshold."
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```



## After: The Code

::: {style="font-size: 0.75em;"}
```{r}
#| eval: false
#| code-block-font-size: 0.6em
nhanes.samp %>% filter(!is.na(HealthGen), !is.na(BPSysAve)) %>% 
ggplot(aes(x = HealthGen, y = BPSysAve, fill = HealthGen)) +
  geom_boxplot(alpha = 0.85, outlier.shape = 21) +
  geom_hline(yintercept = 130, linetype = "dashed",
             color = "grey40", linewidth = 0.8) +
  annotate("text", x = 1, y = 134,
           label = "Hypertension threshold (130 mmHg)",
           hjust = 0, size = 4.5, color = "grey30") +
  scale_fill_manual(
    values = c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00"),
    guide  = "none"
  ) +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 1),
    limits = c(75, 180),
    breaks = seq(80, 180, by = 20)
  ) +
  labs(
    title    = "Systolic blood pressure increases with poorer self-reported health",
    subtitle = "NHANES sample adults (n = 200)",
    x        = "Self-Reported Health Status",
    y        = "Systolic Blood Pressure (mmHg)",
    caption  = "Data: NHANES sample from the oibiostat package."
  ) +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
```
:::



## Key Takeaways

::: {style="font-size: 0.8em;"}
::: {.columns}
::: {.column width="60%"}

**Themes**

- `theme_bw()`, `theme_classic()`, `theme_minimal()` for clean backgrounds
- `theme()` for fine-tuning after the base theme
- `base_size` scales all text proportionally

**Colors**

- Default colors are not colorblind-accessible
- Okabe-Ito: 8 hardcoded hex values, works for all audiences
- `scale_fill_manual()` for filled geoms; `scale_color_manual()` for outlines/points

**Axes and Labels**

- `scale_y_continuous(labels = scales::number_format())` for clean tick labels
- `labs()` is your primary communication tool — use all of it
- Titles should state the finding, not the variable name

**Annotate**

- `geom_hline()` / `geom_vline()` for reference lines
- `annotate("text")` for labels placed in data coordinates
:::

::: {.column width="40%"}
::: {.callout-tip icon="false"}
**The rule:**

Every element of a plot is either earning its place by communicating something, or it is visual noise. Themes, colors, labels, and annotations are how you make every pixel earn its place.
:::
:::
:::
:::



## References

\

::: {style="font-size: 0.8em;"}
::: columns
::: {.column width="50%"}
**Themes and `theme()`**

- [ggplot2 theme documentation](https://ggplot2.tidyverse.org/reference/theme.html)
- [Complete themes in ggplot2](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [ggprism package](https://csdaw.github.io/ggprism/) — Prism theme



\

**Colorblind-accessible palettes**

- [Okabe-Ito palette](https://jfly.uni-koeln.de/color/) — original source
- [Dataviz with R: Color](https://clauswilke.com/dataviz/color-basics.html) — Claus Wilke

\

**Scales and axes**

- [scales package documentation](https://scales.r-lib.org/)
- [ggplot2: Scales](https://ggplot2.tidyverse.org/reference/index.html#scales)

\
:::
::: {.column width="50%"}
**Labels and annotation**

- [ggplot2: annotate()](https://ggplot2.tidyverse.org/reference/annotate.html)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) — Claus Wilke (free online)
- [Data Visualization: A Practical Introducion](https://socviz.co/) — Kieran Healy (free online)

\

**For fun**

- [ggsci package](https://nanx.me/ggsci/) — color palettes inspired by scientific journals, data visualization libraries, science fiction movies, and TV shows.
- [Wes Anderson palettes](https://github.com/karthik/wesanderson) 
:::
:::
:::



