---
title: "BMSC 620 — Homework 4"
subtitle: "Normal, Binomial, and Poisson Distributions"
author: "Your Name"
format: 
  html:
    embed-resources: true
execute:
  echo: true
---

```{r}
#| include: false
#### Global chunk options -----------------------------

knitr::opts_chunk$set(eval = FALSE)

```

::: {.callout-note}
**File naming and submission**

Please use the following file naming convention:

`Lastname_FirstInitial_HW##`

Examples:

- Smith_J_HW04.qmd
- Smith_J_HW04.html

When submitting Homework 4, upload **both**:

- the `.qmd` file
- the rendered `.html` file
:::

## Instructions

- Answer each question clearly and concisely.
- Show your work when calculations are involved.
- Use R to compute probabilities and quantiles when requested.
- You may round final answers to 3 decimal places unless otherwise noted.
- **Do not use continuity correction** for Normal approximations in this assignment.


::: {.callout-tip}
Some code chunks are intentionally incomplete. Fill in the missing values so the code runs correctly.
:::

---

## Part 1: The Normal Distribution

### 1A. Properties of the Normal distribution

Answer briefly (1–2 sentences each).

1. What does it mean for a distribution to be symmetric?
2. What parameters define a Normal distribution?
3. Why is the Normal distribution important in statistics?

**Answers:**

1. 
2. 
3. 

---

## Part 2: Z-scores and Standardization

Suppose adult systolic blood pressure (SBP) is Normally distributed with:

- Mean = 120 mmHg
- Standard deviation = 15 mmHg

### 2A. Calculating z-scores

1. Compute the z-score for an individual with SBP = 135 mmHg.
2. Interpret this z-score in words.

```{r}
# Define parameters
mu <- 120
sigma <- 15
x <- 135

# Compute z-score: z = (x - mu) / sigma
z <- 

z
```

**Answers:**

1. 
2. 

---

## Part 3: Probabilities from the Normal Distribution

Using the same SBP distribution:

### 3A. Probabilities

1. What proportion of adults have SBP above 140 mmHg?
2. What proportion have SBP between 110 and 130 mmHg?

```{r}
# Proportion above 140
pnorm(
  q = ,
  mean = ,
  sd = ,
  lower.tail = 
)

# Proportion between 110 and 130
# P(110 < X < 130) = P(X < 130) - P(X < 110)


```

**Answers:**

1. 
2. 

---

## Part 4: Quantiles and Cutoffs

### 4A. Percentiles

1. What SBP value corresponds to the 90th percentile?
2. Hypertension screening flags the top 10% of SBP values. What is the cutoff?

```{r}
# 90th percentile
qnorm(
  p = ,
  mean = ,
  sd = 
)
```

**Answers:**

1. 
2. 

Interpret the cutoff in one sentence:

---

## Part 5: Binomial Distribution

A new medication has a 30% chance of success for each patient. Suppose 20 patients are treated.

### 5A. Exact Binomial probabilities

1. What is the probability that exactly 8 patients respond?
2. What is the probability that at least 5 patients respond?

```{r}
# Exactly 8 successes
dbinom(
  x = ,
  size = ,
  prob = 
)

# At least 5 successes
# P(X >= 5) = P(X > 4) = 1 - P(X <= 4)
pbinom(
  q = ,
  size = ,
  prob = ,
  lower.tail = 
)
```

**Answers:**

1. 
2. 

---

## Part 6: Normal Approximation to the Binomial

Suppose a manufacturing process has a defect rate of 10%. Out of 200 items produced:

### 6A. Check conditions

1. State the mean and standard deviation of the Binomial distribution.
2. Verify whether a Normal approximation is reasonable.

```{r}
n <- 200
p <- 0.10

# Mean of binomial: mu = n * p
mu_b <- 

# SD of binomial: sd = sqrt(n * p * (1 - p))
sd_b <- 

mu_b
sd_b

# Check conditions: np >= 10 and n(1-p) >= 10


```

**Answers:**

1. 
2. 

### 6B. Approximate probability

Use a Normal approximation **without continuity correction**.

What is the approximate probability that more than 30 items are defective?

```{r}
# P(X > 30) using Normal approximation
pnorm(
  q = ,
  mean = ,
  sd = ,
  lower.tail = 
)
```

**Answer:**

Briefly comment on whether this approximation seems reasonable:

---

## Part 7: Poisson Distribution

Emergency room arrivals follow a Poisson distribution with an average rate of 3 arrivals per hour.

### 7A. Probabilities

1. What is the probability of exactly 2 arrivals in an hour?
2. What is the probability of at least 5 arrivals in an hour?

```{r}
# Exactly 2 arrivals
dpois(
  x = ,
  lambda = 
)

# At least 5 arrivals: P(X >= 5) = 1 - P(X <= 4)
ppois(
  q = ,
  lambda = ,
  lower.tail = 
)
```

**Answers:**

1. 
2. 

---

## Part 8: Choosing the Right Distribution

For each scenario, state which distribution (**Normal**, **Binomial**, or **Poisson**) is most appropriate and briefly explain why.

1. Number of typos on a randomly selected manuscript page
2. Proportion of voters in a sample who support a candidate
3. Adult heights in a large population

**Answers:**

1. 
2. 
3. 

---

## Optional (Not Graded): Reflection

In 2–3 sentences, describe one situation where using the wrong distribution could lead to misleading conclusions.

**Answer:**

---

## Submission checklist

Before submitting, make sure you have:

- [ ] Answered all required questions
- [ ] Included R output where requested
- [ ] Written brief interpretations in words when asked
- [ ] Rendered your HTML successfully
- [ ] Uploaded both .qmd and .html files
