---
title: "Final Exam Review"
subtitle: "Cumulative: Lessons 1–18, Emphasis on Lessons 10–19"
author: "Emile Latour"
date: "`r library(here); source(here('class_dates.R')); w11d1`"
date-format: long
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v5.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | Final Exam Review"
    html-math-method: mathjax
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(here)
library(tidyverse)
library(oibiostat)
library(broom)
library(rstatix)
library(janitor)
library(pwr)
library(laviz)
library(lamisc)

theme_set(laviz::theme_minimal_white(grid = "none",
                                     axis = "xy",
                                     base_size = 16))

data(nhanes.samp)
data(prevend.samp)

set.seed(620)
```

## Today's Agenda

- Exam logistics and structure
- First-half foundations (what's on the exam from Lessons 1–9)
- Second-half concepts — the main focus:
  - Data exploration
  - Paired t-test
  - ANOVA and post-hoc testing
  - Categorical data analysis
  - Nonparametric tests
  - Correlation and simple linear regression
  - Power, sample size, and synthesis
- Choosing the right test
- Your questions

\

**The final exam opens TODAY at 3:00 PM and closes Friday, March 20 at 5:00 PM**

# Exam Logistics

## Exam Details

**Format:**

- 8 parts, 150 points total
- Designed for 4–6 hours completion
- Mix of conceptual questions, R code, and interpretation

\

**Resources allowed:**

- ✅ Course materials (slides, homework, textbook)
- ✅ R documentation and help files (`?t.test`, `?pnorm`, etc.)
- ✅ Your own notes
- ❌ Other students
- ❌ AI assistants (ChatGPT, Claude, etc.)
- ❌ Online forums or external help

\

**Submit both `.qmd` and `.html` files — naming: `Lastname_FirstInitial_Final`**

## What's Covered

\

\

::::: columns
::: {.column width="48%"}

**Part 1:** Foundations — first half (15 pts)

**Part 2:** Exploring the data (15 pts)

**Part 3:** Paired t-test (20 pts)

**Part 4:** ANOVA and post-hoc (20 pts)

:::

::: {.column width="48%"}

**Part 5:** Categorical data (20 pts)

**Part 6:** Nonparametric tests (15 pts)

**Part 7:** Correlation and SLR (25 pts)

**Part 8:** Power, sample size, synthesis (20 pts)

:::
:::::

\

**Key dataset:** `diabetes_trial.csv`

- Simulated RCT, 180 patients, 3 treatment arms (Metformin, Lifestyle, Combined), 6-month follow-up
- Read the variable descriptions carefully before starting

## General Tips

\

::: columns
::: {.column width="50%"}

**Before you start:**

- Set up your workspace: load packages, read in data
- Make sure code chunks evaluate 
- Check that `here("data", "your_file.csv")` finds your file

\

**While working:**

- Interpret everything in complete sentences — reference the numbers
- Use all 6 steps for hypothesis tests
- Use comments in your R code to show your thinking
- **Save often**

:::

::: {.column width="5%"}
:::

::: {.column width="45%"}

**Before submitting:**

- Render and confirm no errors
- Verify both `.qmd` and `.html` are submitted
- Check file naming format

\

<!-- ::: {.callout-warning icon="false"} -->
<!-- ### The #1 mistake -->

<!-- Running code that *works* but never rendering the document. A file that won't knit cannot be graded. Render early and often. -->
<!-- ::: -->

:::
:::

# Part 1: First-Half Foundations

## What from the first half is on the exam?

\

Part 1 (15 points) tests three targeted areas from Lessons 1–9:

\

::: columns
::: {.column width="50%"}
**1.1 — Study design (5 pts)**

- Experiment vs. observational study
- Why randomization matters
- When can we claim causation?

**1.2 — Bayes' theorem / PPV (5 pts)**

- Given sensitivity, specificity, prevalence → calculate PPV
- Interpret PPV in context
:::
::: {.column width="50%"}
**1.3 — Normal distributions and CLT (5 pts)**

- `pnorm()` for an individual value
- `pnorm()` for a sample mean (using $SE = \frac{\sigma}{\sqrt{n}}$)
:::
:::




\

::: {.callout-tip icon="false"}
Your **midterm review slides** cover all of this in detail — a quick re-read is the best preparation for Part 1.
:::

## Study design: the essentials

\

::::: columns
::: {.column width="48%"}

**Experiment:**

- Researchers **assign** treatments (random assignment)
- Controls for confounding
- Can establish **causation**

**Observational study:**

- Researchers observe without interfering
- Cannot control for all confounders
- Can only show **association**

:::

::: {.column width="48%"}

::: {.callout-important icon="false"}
### Causation checklist

A study can claim causation **only if:**

1. It is a randomized experiment
2. Random assignment was used
3. Other explanations have been ruled out

**Key question:** Was treatment *assigned* by the researchers, or did subjects choose/arrive with it?
:::

:::
:::::

## Bayes' Theorem: PPV setup

Given sensitivity, specificity, and prevalence:

$$PPV = P(\text{Disease} \mid \text{Test}+) = \frac{\text{sensitivity} \times \text{prevalence}}{\text{sensitivity} \times \text{prevalence} + (1 - \text{specificity}) \times (1 - \text{prevalence})}$$

\

**Worked example:** A TB screening test has sensitivity = 0.79, specificity = 0.87, applied in a population with 6% prevalence.

```{r}
sensitivity <- 0.79
specificity <- 0.87
prevalence  <- 0.06

ppv <- (sensitivity * prevalence) /
  (sensitivity * prevalence + (1 - specificity) * (1 - prevalence))
ppv
```

::: {.callout-tip icon="false"}
**Interpreting PPV:** "Of all patients who test positive, ___% actually have the condition."

A high PPV → test is useful for **confirming** disease.
A low PPV → test is better for **screening** (ruling out).
:::

## Normal distributions and CLT: two scenarios

\

::::: columns
::: {.column width="48%"}

**Individual observation:** $X \sim N(\mu, \sigma^2)$

```{r}
# LDL cholesterol: mu = 130, sd = 35
# P(LDL > 160) — "borderline high"

pnorm(160, mean = 130, sd = 35,
      lower.tail = FALSE)
```

\

The SD here is $\sigma$ — variability of **individuals**

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**Sample mean:** $\bar{X} \sim N\left(\mu, \frac{\sigma^2}{n}\right)$

```{r}
# for n = 25, same population
# P(Xbar > 140) 

se <- 35 / sqrt(25)
pnorm(140, mean = 130, sd = se,
      lower.tail = FALSE)
```

\

The variability here is measure by $SE = \sigma/\sqrt{n}$ — variability of **sample means**

:::
:::::

\

::: {.callout-important icon="false"}
### The key distinction

When the question says "a random individual" → use $\sigma$.

When the question says "a sample of $n$" → use $SE = \sigma/\sqrt{n}$.

:::

# Part 2: Exploring the Data

## Descriptive statistics by group

```{r}
#| eval: false

# Overall summary of a continuous variable
nhanes.samp %>%
  summarise(
    n      = sum(!is.na(Pulse)),
    mean   = mean(Pulse, na.rm = TRUE),
    sd     = sd(Pulse, na.rm = TRUE),
    median = median(Pulse, na.rm = TRUE)
  )

# By group
nhanes.samp %>%
  group_by(Gender) %>%
  summarise(
    n    = sum(!is.na(Pulse)),
    mean = mean(Pulse, na.rm = TRUE),
    sd   = sd(Pulse, na.rm = TRUE)
  )

# Or with rstatix
nhanes.samp %>%
  group_by(Gender) %>%
  get_summary_stats(Pulse, type = "mean_sd")
```

## Visualization: histograms and boxplots

```{r}
#| eval: false

# Histogram — check shape, skewness, outliers
ggplot(nhanes.samp, 
       aes(x = BPSys1)) +
  geom_histogram(bins = 30, 
                 color = "white") +
  labs(x = "Systolic BP (mmHg)", y = "Count",
       title = "Distribution of systolic blood pressure")

# Boxplots by group with individual points
ggplot(nhanes.samp, 
       aes(x = Gender, y = Pulse, fill = Gender)) +
  geom_boxplot(alpha = 0.6, 
               width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  labs(x = "Gender", y = "Pulse (bpm)",
       title = "Pulse rate by gender")
```

\

::: {.callout-tip icon="false"}
`geom_jitter()` is layered **after** `geom_boxplot()` so points appear on top. Always include `width = 0.2` or similar to prevent overplotting.
:::

## Frequency tables with `tabyl()`

```{r}
#| eval: false

# One-way table
nhanes.samp %>%
  tabyl(SmokeNow)

# Two-way table with row percentages
nhanes.samp %>%
  tabyl(PhysActive, SmokeNow) %>%
  adorn_totals(where = "row") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()
```

\

::: {.callout-tip icon="false"}
`adorn_percentages("row")` gives **row percentages** — the proportion *within* each row group. `adorn_ns()` keeps the raw counts alongside so both $n$ and % are visible.
:::

# Part 3: Paired t-test

## When is a paired design appropriate?

\

::::: columns
::: {.column width="48%"}

**Paired (dependent) design:**

- Same subject measured **twice** (before/after)
- Matched pairs (sibling controls, eye studies)
- The pairing controls for between-subject variability

**Recognize it by:** 

- the same ID appears in both measurements
- there is a natural link between observations

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-important icon="false"}
### Why pairing matters

- If you ignore the pairing and treat the two measurements as independent groups, you **throw away information** about within-subject change. 
- The paired test uses differences $d_i = Y_{i,\text{post}} - Y_{i,\text{pre}}$ and is equivalent to a one-sample t-test on those differences.
- Ignoring pairing → larger SE → less power → more likely to miss a real effect.
:::

:::
:::::

## Paired t-test: the workflow

\

**Step 1: Create the difference variable**

```{r}
#| eval: false

# Example: SBP before and after a 12-week exercise intervention
bp_data <- bp_data %>%
  mutate(sbp_change = sbp_pre - sbp_post)  # positive = decrease in BP
```

\

**Step 2: Run the test**

```{r}
#| eval: false

# Option A: one-sample t-test on differences
t.test(bp_data$sbp_change, mu = 0, alternative = "two.sided")

# Option B: paired t-test directly
t.test(bp_data$sbp_pre, bp_data$sbp_post,
       paired = TRUE, alternative = "two.sided")
```

\

::: {.callout-tip icon="false"}
Both give identical results. Option A is often clearer for applying the six steps — you are explicitly testing whether the mean difference equals zero.
:::

## Paired t-test: the six steps

\

**Step 1:** $H_0: \mu_d = 0$ (no mean change) vs. $H_A: \mu_d > 0$ (BP decreased, if directional)

**Step 2:** $\alpha = 0.05$

**Step 3: Assumptions**

- Independence: one pair per subject ✓ (study design)
- Normality of differences: large $n$ → CLT; small $n$ → check histogram or QQ plot of differences

**Step 4:** $t = \dfrac{\bar{d} - 0}{s_d / \sqrt{n}}$ — reported as `t` in `t.test()` output

**Step 5:** p-value from `t.test()` output; $df = n - 1$

**Step 6:** Reject if p < 0.05. Report in context:

> "There was a statistically significant decrease in SBP from baseline to 12 weeks (t = \_\_\_, df = \_\_\_, p = \_\_\_). On average, SBP decreased by \_\_\_ mmHg (95% CI: \_\_\_ to \_\_\_)."

# Part 4: ANOVA and Post-Hoc Testing

## When to use one-way ANOVA

::: {.callout-note icon="false"}
### ANOVA: comparing means across 3+ groups

Use one-way ANOVA when:

- **Outcome:** continuous
- **Predictor:** categorical with **3 or more groups**
- **Question:** Do any of the group means differ?
:::

\

**Why not just run multiple t-tests?**

- Each t-test has a 5% Type I error rate. 
- With 3 pairwise comparisons, the family-wise error rate inflates to ~14% — you're much more likely to find a spurious significant result by chance alone.
- **ANOVA tests all groups simultaneously**, preserving the overall error rate at $\alpha$.

## ANOVA assumptions (1/3)

1. **Independence:** one observation per subject (study design) ✓
2. **Normality:** approximately normal within each group
3. **Equal variance:** SD ratio (max SD / min SD) should be < 2

\

```{r}
#| eval: true

# Create 3 age groups, then check group-level stats
nhanes_age <- nhanes.samp %>%
  mutate(age_group = cut(Age,
                         breaks = c(0, 35, 55, 100),
                         labels = c("Young", "Middle", "Older"))) %>%
  filter(!is.na(age_group), !is.na(BPSys1))

```

## ANOVA assumptions (2/3): normality

**Check:** distributions approximately normal within each group

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

ggplot(nhanes_age, aes(x = BPSys1)) +
  geom_histogram(bins = 20, color = "white") +
  facet_wrap(~ age_group) +
  labs(x = "Systolic BP (mmHg)", y = "Count",
       title = "Distribution of systolic BP by age group")
```

::: {.callout-tip icon="false"}
With reasonably large groups, moderate departures from normality are acceptable — the CLT helps. Look for severe skew or extreme outliers.
:::

## ANOVA assumptions (3/3): equal variance

**Check:** SD ratio (max SD / min SD) should be < 2

```{r}
nhanes_age %>%
  group_by(age_group) %>%
  summarise(n    = n(),
            mean = mean(BPSys1),
            sd   = sd(BPSys1)) %>%
  mutate(sd_ratio = max(sd) / min(sd))
```

::: {.callout-tip icon="false"}
SD ratio < 2 → equal variance assumption is met ✓

If the ratio is borderline (say, 1.8–2.5), note the concern but proceed — ANOVA is fairly robust to mild violations when group sizes are similar.
:::

## ANOVA: the six steps + R code

\

**Step 1:** $H_0: \mu_1 = \mu_2 = \mu_3$ vs. $H_A:$ at least one group mean differs

\

**Steps 2–3:** $\alpha = 0.05$; check assumptions ✓

\

**Steps 4–5: Fit and summarize**

```{r}
#| eval: false

anova_fit <- aov(BPSys1 ~ age_group, data = nhanes_age)

# Traditional output
summary(anova_fit)

# Tidy output
broom::tidy(anova_fit)
```


```{r}
#| include: false
anova_fit <- aov(BPSys1 ~ age_group, data = nhanes_age)

# Tidy output
aov_res <- broom::tidy(anova_fit)

```

\

**Step 6:** Report F-statistic with **both** df values and p-value

> "There was a significant difference in mean systolic blood pressure between at least two age groups (F(`r aov_res$df[1]`, `r aov_res$df[2]`) = \_\_\_, p = \_\_\_)."

## Post-hoc testing: Tukey's HSD

If ANOVA rejects $H_0$, use Tukey HSD to identify **which pairs** differ:

```{r}
#| eval: false

rstatix::tukey_hsd(anova_fit)
```

\

**Reading the output:**

- `estimate` = difference in group means
- `p.adj` = adjusted p-value (controls family-wise error rate)
- Report pairs where `p.adj < 0.05`

\

::: {.callout-important icon="false"}
### Writing a results paragraph

Include: descriptive stats for each group + F-statistic + p-value + specific post-hoc comparisons.

> "Mean [outcome] was \_\_\_ (SD = \_\_\_) in Group A, \_\_\_ (SD = \_\_\_) in Group B, and \_\_\_ (SD = \_\_\_) in Group C. A one-way ANOVA indicated a significant difference between group means (F(df1, df2) = \_\_\_, p = \_\_\_). Tukey post-hoc tests revealed that [Group A vs. Group C: mean diff = \_\_\_, p = \_\_\_] differed significantly, while [other comparisons] did not."
:::

# Part 5: Categorical Data Analysis

## Chi-squared test for independence

\

**When to use:** Both variables are **categorical**, testing whether they are associated

\

```{r}
#| eval: false

# Step 1: Build the contingency table
tab <- nhanes.samp %>%
  filter(!is.na(SmokeNow), !is.na(PhysActive)) %>%
  tabyl(PhysActive, SmokeNow)

# Step 2: Check expected counts — all should be ≥ 5
chisq.test(tab)$expected

# Step 3: Run the test
chisq.test(tab)
```

\

**Hypotheses (in words):**

- $H_0$: There is no association between physical activity and smoking status
- $H_A$: There is an association between physical activity and smoking status

## Two-proportion z-test

\

**When to use:** Exactly two groups, binary outcome

\

```{r}
#| eval: false

# Compute successes and group sizes
counts <- nhanes.samp %>%
  filter(!is.na(SmokeNow)) %>%
  group_by(Gender) %>%
  summarise(
    success = sum(SmokeNow == "Yes", na.rm = TRUE),
    n       = sum(!is.na(SmokeNow))
  )

# Two-proportion test
prop.test(x = counts$success,
          n = counts$n,
          alternative = "two.sided")
```

\

**Hypotheses:**

- $H_0$: $p_1 = p_2$
- $H_A$: $p_1 \ne p_2$

## Chi-squared vs. two-proportion test

\

| | Chi-squared | Two-proportion (`prop.test`) |
|---|---|---|
| **Groups** | 2 or more | Exactly 2 |
| **Outcome** | Categorical (any # levels) | Binary (success/failure) |
| **Hypotheses** | Independence vs. association | $p_1 = p_2$ vs. $p_1 \neq p_2$ |
| **Output** | $\chi^2$, df, p-value | z (or $\chi^2$), p-value, **CI for difference** |

\

::: {.callout-tip icon="false"}
For exactly **2 groups and a binary outcome**, both tests give equivalent p-values. Prefer `prop.test()` when you want a **confidence interval for the difference in proportions** — that CI gives more information than the p-value alone.
:::

# Part 6: Nonparametric Tests

## When to go nonparametric

::: {.callout-note icon="false"}
### The decision is about the data, not just sample size

Use a nonparametric test when:

- The distribution is **heavily skewed** or has extreme outliers
- The outcome is **ordinal** (pain scales, severity scores)
- Sample sizes are small **and** normality is implausible
:::

\

**How to assess:** plot the data first (histogram or boxplot); check the SD ratio across groups if comparing 3+ groups

\

::: {.callout-warning icon="false"}
Small $n$ alone does not automatically mean nonparametric. If the data look approximately normal, the parametric test is valid and more powerful. The criterion is whether the **assumptions** are plausibly met.
:::

## Three nonparametric tests: the map

\

| Parametric test | Nonparametric alternative | R function |
|---|---|---|
| One-sample / paired t-test | Wilcoxon signed-rank | `wilcox.test(x, mu = 0)` |
| Two-sample t-test | Wilcoxon rank-sum (Mann-Whitney) | `wilcox.test(x, y)` |
| One-way ANOVA | Kruskal-Wallis | `kruskal.test()` |

\

**Hypotheses (state correctly!):**

- $H_0$: The distributions of [outcome] are **identical** across groups
- $H_A$: At least one group has a different distribution

::: {.callout-important icon="false"}
Do **not** state hypotheses in terms of medians being equal — this is only valid under additional assumptions. "Distributions are identical" is the correct framing.
:::

## Kruskal-Wallis + pairwise follow-up

```{r}
#| eval: false

# Kruskal-Wallis test
kw_result <- kruskal.test(Pulse ~ age_group, data = nhanes_age)

kw_result %>%
  broom::tidy()

# Pairwise Wilcoxon with Bonferroni correction (if K-W is significant)
pairwise.wilcox.test(nhanes_age$Pulse, nhanes_age$age_group,
                     p.adjust.method = "bonferroni") %>% 
  broom::tidy()
```

\

::: {.callout-tip icon="false"}
<!-- Use **FDR** (Benjamini-Hochberg) correction for multiple comparisons — it controls the **false discovery rate** and is less conservative than Bonferroni when you have several comparisons. -->
There is no nonparametric equivalent of Tukey HSD. Bonferroni correction is a straightforward choice — it controls the family-wise error rate across all pairwise comparisons. With only 3 pairs here, the penalty for using Bonferroni is small.
:::

# Part 7: Correlation and Simple Linear Regression

## This section is 25 points — the largest on the exam

\

::: {.callout-important icon="false"}
### No homework covered this material

Part 7 is the only section that was not practiced in a homework assignment. The exam provides extra guidance (hints in code chunks). Read the questions carefully and follow the steps.

The material maps directly to the Lesson 21 slides — a review of those slides before the exam is time well spent.
:::

\

**What Part 7 tests:**

- Scatterplot + regression line
- Pearson and Spearman correlation: `cor()`
- Fitting `lm()`, interpreting `tidy()` output
- LINE assumption checks with `augment()` + diagnostic plots
- $R^2$ from `glance()`
- Prediction vs. confidence interval: `predict()`

## Scatterplot and correlation

Using `prevend.samp`: does cognitive function (`RFFT` score) decline with age?

```{r}
# Scatterplot with regression line
ggplot(prevend.samp, aes(x = Age, y = RFFT)) +
  geom_point(alpha = 0.4, color = "#0072B2", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#D55E00") +
  labs(x = "Age (years)", y = "RFFT score",
       title = "Cognitive function vs. age (prevend.samp)")
```

## Correlation coefficients

```{r}
# Pearson: strength of LINEAR association
cor(prevend.samp$Age, prevend.samp$RFFT,
    use = "complete.obs")

# Spearman: rank-based, robust to skew/outliers
cor(prevend.samp$Age, prevend.samp$RFFT,
    method = "spearman",
    use = "complete.obs")
```

\

**Describing the relationship:** direction (positive/negative), form (linear/curved), strength (weak/moderate/strong), unusual observations

::: {.callout-tip icon="false"}
The Pearson and Spearman values are usually close when the relationship is approximately linear without extreme outliers. A large gap suggests outliers or nonlinearity are influencing the Pearson $r$.
:::

## Fitting the regression model

```{r}
# Fit
rfft_model <- lm(RFFT ~ Age, data = prevend.samp)

# Coefficients, SE, t, p, CI
tidy(rfft_model, conf.int = TRUE)
```

\

**Interpreting the slope ($\hat{\beta}_1$):**

> "For every 1-year increase in age, RFFT score is expected to change, on average, by $\hat{\beta}_1$ points (95% CI: \_\_\_ to \_\_\_)."

\

**Interpreting the intercept ($\hat{\beta}_0$):** Often not meaningful — only interpret when $X = 0$ is realistic in context.

## The LINE conditions

| Letter | Condition | How to check |
|:---:|---|---|
| **L** | **L**inearity — relationship is linear | Scatterplot; residual plot |
| **I** | **I**ndependence — observations independent | Study design |
| **N** | **N**ormality — residuals approximately normal | QQ plot of residuals |
| **E** | **E**qual variance — residuals have constant spread | Residual plot |

: {tbl-colwidths="\[13, 40, 20\]"}

\

```{r}
#| eval: false

aug <- augment(rfft_model)

# Residual plot (checks L and E)
ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted values", y = "Residuals", title = "Residual plot")

# QQ plot (checks N)
ggplot(aug, aes(sample = .resid)) +
  stat_qq(alpha = 0.4) + stat_qq_line() +
  labs(title = "QQ plot of residuals")
```

## Diagnostic plots for rfft_model

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-pos: center

aug <- augment(rfft_model)

# Residual plot (checks L and E)
a <- ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted values", y = "Residuals", title = "Residual plot")

# QQ plot (checks N)
b <- ggplot(aug, aes(sample = .resid)) +
  stat_qq(alpha = 0.4) + stat_qq_line() +
  labs(title = "QQ plot of residuals")

library(patchwork)
a + b
```


## Reading the diagnostic plots

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Residual plot (L and E)

**Good:** Points scatter randomly around 0; flat smoother; roughly equal spread

**Red flags:**

- Curve → linearity violated (L)
- Funnel → equal variance violated (E)
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### QQ plot (N)

**Good:** Points track the diagonal reference line

**Red flags:**

- S-shape → skewed residuals
- Heavy tails → points deviate at both ends

**Note:** With large $n$, the CLT makes moderate departures from normality less of a concern. L and E are more critical for inference.
:::

:::
:::::

## $R^2$ and sigma

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### $R^2$ interpretation

"Age explains ___% of the variability in RFFT score in this sample."

$R^2$ ranges from 0 (model explains nothing) to 1 (perfect fit).

For SLR: $R^2 = r^2$ (square of Pearson correlation)
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### sigma

`sigma` = residual standard deviation = typical distance of an individual observation from the fitted line, in **units of $Y$**.

This captures individual-level scatter. It is much larger than `std.error` for the slope, and it is why prediction intervals are so much wider than CIs for mean response.
:::

:::
:::::

\

```{r}
# Model-level summary: R², sigma, n
glance(rfft_model) %>% select(r.squared, sigma, nobs)
```

::: {.callout-tip icon="false"}
`std.error` (from `tidy()`) tells you how well you estimated the slope. `sigma` (from `glance()`) tells you how much individuals vary around the line. These measure completely different things.
:::




## Inference for the slope: six steps

\

**Step 1:** $H_0: \beta_1 = 0$ (age tells us nothing about RFFT) vs. $H_A: \beta_1 \neq 0$

\

**Step 2:** $\alpha = 0.05$

\

**Step 3:** Check LINE conditions ✓

\

**Step 4:** $t = \hat{\beta}_1 / SE_{\hat{\beta}_1}$ — read from `tidy()` output, `statistic` column

\

**Step 5:** p-value from `tidy()` output, `p.value` column; $df = n - 2$

\

**Step 6:** Conclusion in context:

> "There is strong evidence of a negative linear association between age and RFFT score (t = \_\_\_, df = \_\_\_, p = \_\_\_). For every 1-year increase in age, RFFT score is expected to decrease by \_\_\_ points (95% CI: \_\_\_ to \_\_\_)."

## CI for mean response vs. prediction interval

| | Question | Accounts for |
|---|---|---|
| **CI for mean response** | Average RFFT for **all people** aged 60? | Uncertainty about where the line is |
| **Prediction interval** | RFFT for **this specific new patient** aged 60? | Uncertainty about the line **+** individual variation ($\hat{\sigma}$) |

: {tbl-colwidths="\[20, 40, 30\]"}

\

```{r}
#| eval: false

new_data <- data.frame(Age = c(45, 60, 75))

# CI for mean response
predict(rfft_model, newdata = new_data, interval = "confidence")

# Prediction interval for a new individual
predict(rfft_model, newdata = new_data, interval = "prediction")
```

\

::: {.callout-important icon="false"}
Prediction intervals are **always wider** than CIs for mean response. More data narrows the CI (better-estimated line) but cannot eliminate individual variation — that scatter is a property of the population, not of your sample size.
:::

# Part 8: Power, Sample Size, and Synthesis

## Power analysis with `pwr.t.test()`

**Cohen's $d$:** effect size for comparing two means

$$d = \frac{\mu_1 - \mu_2}{\sigma_{\text{pooled}}}$$

\

**Example:** detect a 5 mmHg difference in SBP, pooled SD = 10 → $d = 0.5$

::::: columns
::: {.column width="48%"}
**Solve for $n$:**
```{r}
#| eval: false
pwr.t.test(n           = NULL,
           d           = 5 / 10,
           sig.level   = 0.05,
           power       = 0.80,
           type        = "two.sample",
           alternative = "two.sided")
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
**Solve for power** with n = 50 per group:
```{r}
#| eval: false
pwr.t.test(n           = 50,
           d           = 5 / 10,
           sig.level   = 0.05,
           power       = NULL,
           type        = "two.sample",
           alternative = "two.sided")
```
:::
:::::

::: {.callout-tip icon="false"}
Always **round up** the required $n$ — rounding down means you're slightly underpowered.
:::

## Sample size and what it affects

\

When you increase $n$:

| | Effect |
|---|---|
| **ANOVA F-statistic** | Increases (more power to detect group differences) |
| **Post-hoc p-values** | Decrease (more power per pairwise comparison) |
| **Width of CIs** | Narrower (smaller SE → tighter interval) |

: {tbl-colwidths="\[20, 40\]"}

\

::: {.callout-important icon="false"}
### Peeking at data before choosing the test

- Running ANOVA, seeing which group performed best, then running a t-test comparing that group vs. the others is **data-driven test selection** — it inflates the Type I error rate because you used the data twice. 
- The right approach: specify your comparisons **before** looking at the data.
:::

## Choosing the right test: the synthesis table

\

| Research question | Groups | Outcome | Test | Key R function |
|---|:---:|:---:|---|---|
| Did the outcome change? (same subject, 2 timepoints) | 1 (differences) | Continuous | Paired t-test | `t.test(d, mu = 0)` |
| Does mean outcome differ across 3+ groups? | 3+ | Continuous, normal | One-way ANOVA | `aov()` |
| Does a skewed outcome differ across 3+ groups? | 3+ | Continuous, skewed | Kruskal-Wallis | `kruskal.test()` |
| Is a categorical variable associated with a binary outcome? | 3+ | Binary | Chi-squared | `chisq.test()` |
| Does a binary outcome rate differ between 2 groups? | 2 | Binary | Two-proportion z-test | `prop.test()` |
| Is there a linear association between two continuous variables? | — | 2 continuous | SLR / slope test | `lm()` + `tidy()` |

: {tbl-colwidths="\[35, 15, 15, 20, 25\]"}

## CI and hypothesis test: two sides of the same coin

::: {.callout-note icon="false"}
### A recurring exam theme

For a two-sided test at $\alpha = 0.05$:

- If the **95% CI excludes** the null value → p-value < 0.05 → reject $H_0$
- If the **95% CI includes** the null value → p-value ≥ 0.05 → fail to reject $H_0$
:::

\

**What the CI adds beyond the p-value:**

- The **direction and magnitude** of the effect (not just "significant or not")
- A range of **plausible values** for the true population parameter
- Clinical interpretability — a statistically significant effect may be too small to matter in practice

\

**Good examples:** any test where you computed both a CI and a p-value. Ask: what does the CI tell us about the *size* of the effect that the p-value alone cannot?

# Reminders and Good Luck

## Common mistakes to avoid

\

1. **Ignoring paired structure** — running an independent t-test on before/after data
2. **Forgetting to check assumptions** — especially ANOVA's SD ratio and chi-squared expected counts ≥ 5
3. **Misframing nonparametric hypotheses** — say "distributions are identical," not "medians are equal"
4. **Confusing $\sigma$, $s$, $SE$, and `sigma` from `glance()`** — know what each measures
5. **Using `lower.tail = TRUE/FALSE` incorrectly** — sketch the picture first
6. **Not rendering before submitting** — code that works in a chunk but breaks the render is ungraded

\


## You are ready

\

- ✅ You have worked through 8 homework assignments with increasingly complex analyses
- ✅ You have practiced the six-step framework across many test types
- ✅ You have hands-on experience with every R function on this exam
- ✅ Part 7 has hints — you are not going in cold on SLR

\

**The exam opens at 3:00 PM today. You have until Friday at 5:00 PM.**

\

Read carefully, interpret in context, save frequently, and trust your preparation.

**Good luck!**

# Questions?

## Your Turn

\

**What would you like to go over?**

- Concepts that feel unclear?
- R functions you want to see again?
- Specific parts of the exam structure?

\

This is your time — ask anything.
