---
title: "BMSC 620 — Homework 8"
subtitle: "ANOVA and Nonparametric Tests"
author: "Your Name"
format: 
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
  eval: true
---


::: {.callout-note}
**File naming and submission**

Please use the following file naming convention:

`Lastname_FirstInitial_HW08`

Examples:

- Smith_J_HW08.qmd
- Smith_J_HW08.html

When submitting Homework 8, upload **both**:

- the `.qmd` file
- the rendered `.html` file
:::

## Instructions

- Answer each question clearly and concisely.
- Show your work using R code chunks.
- **Rounding:** Round summary statistics to 1 decimal place.
  Round *p*-values to 3 decimal places (or report as *p* < 0.001 if very small).
- This homework uses an **external data file** — see Part 0 below before writing
  any code.
- Code chunks marked `# Your code here` are intentionally blank. You are
  expected to write the code yourself. Hints are provided to guide you.

::: {.callout-note}
**A note on this homework**

This assignment covers two closely related topics. Parts 1–2 focus on ANOVA
and post-hoc testing — extending the two-sample *t*-test to three or more
groups. Parts 3–5 introduce nonparametric alternatives that you can use when
parametric assumptions break down. The dataset is designed so that one
outcome variable meets ANOVA assumptions nicely, while another does not —
giving you practice deciding which approach is appropriate.
:::

---

## Part 0: Project Setup

**Before writing any code, do the following:**

1. Make sure you are working inside your BMSC 620 R Project.
2. Download the data file from OneDrive and place it inside your
   project's `data/` folder:
   - `knee_rehab.csv`
3. Your folder structure should look like this:

```
BMSC620/
├── BMSC620.Rproj
├── data/
│   ├── knee_rehab.csv
└── homework/
    └── Smith_J_HW08.qmd    ← your file goes here
```

Load the packages you will need:

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(broom)
library(rstatix)
library(janitor)
```

---

## Background

The `knee_rehab.csv` dataset comes from a simulated randomized controlled
trial of **150 patients** recovering from total knee replacement surgery.
Patients were randomized equally to one of three rehabilitation protocols:

- **Standard**: Traditional physical therapy (clinic-based exercises)
- **Aquatic**: Water-based rehabilitation (pool therapy)
- **Accelerated**: Intensive early mobilization protocol

The dataset contains the following variables:

| Variable | Description |
|---|---|
| `patient_id` | Unique patient identifier |
| `protocol` | Rehabilitation protocol (Standard, Aquatic, Accelerated) |
| `age` | Patient age (years) |
| `sex` | Patient sex (Female, Male) |
| `rom_improvement` | Range of motion improvement in degrees at 12 weeks |
| `pain_score_12wk` | Self-reported pain score at 12 weeks (0–10 scale) |
| `tug_before` | Timed Up-and-Go test (seconds) before rehabilitation |
| `tug_after` | Timed Up-and-Go test (seconds) after 12 weeks of rehabilitation |

> *Data simulated for educational purposes. Values do not represent real
> patient data.*

---

## Part 1: One-Way ANOVA

### 1A. Load and explore the data

Import `knee_rehab.csv` and assign it to an object called `knee`. Use
`glimpse()` to inspect the structure.

```{r}
#| label: load-knee

# Your code here

```

Compute summary statistics (sample size, mean, and SD) for
`rom_improvement` by `protocol`.

::: {.callout-tip}
Try `group_by(protocol) %>% summarise(n = n(), mean = mean(rom_improvement), sd = sd(rom_improvement))`.
:::

```{r}
#| label: rom-summary

# Your code here

```

Which group had the highest mean ROM improvement? The lowest?

**Answer:**

---

### 1B. Visualize the data

Create side-by-side boxplots of `rom_improvement` by `protocol`. Overlay
individual data points using `geom_jitter()`.

::: {.callout-tip}
Use `geom_boxplot(outlier.shape = NA)` so the jittered points don't
duplicate the outlier markers.
:::

```{r}
#| label: rom-boxplot

# Your code here

```

Based on the boxplots, do the three groups appear to have similar
spread? Does the data within each group look approximately symmetric?

**Answer:**

---

### 1C. State the hypotheses

We want to test whether mean ROM improvement differs across the three
rehabilitation protocols.

State the null and alternative hypotheses in both **symbols** and **words**.

**Answer:**

$H_0$:

$H_A$:

---

### 1D. Check ANOVA assumptions

Before running the ANOVA, check the key assumptions:

1. **Equal variance:** Compute the ratio of the largest to smallest
   within-group SD. Is the ratio less than 2?
2. **Sample size / normality:** Are there at least 30 observations per
   group?

::: {.callout-tip}
You already computed the SDs in 1A. Use those to calculate the max/min
ratio.
:::

```{r}
#| label: check-anova-assumptions

# Your code here

```

Are the ANOVA assumptions reasonably met? Briefly explain.

**Answer:**

---

### 1E. Run the one-way ANOVA

Fit the ANOVA model using `aov()` and display the ANOVA table with
`summary()`. Also display the tidy output using `broom::tidy()`.

```{r}
#| label: run-anova

# Your code here

```

Report the F-statistic, degrees of freedom, and p-value.

**Answer:**

---

### 1F. Interpret the ANOVA result

Write a 2–3 sentence conclusion in context. Be sure to:

- State whether you reject or fail to reject $H_0$ at $\alpha = 0.05$
- State what the result means about ROM improvement across protocols
- Report the F-statistic with both df values and the p-value

**Answer:**

---

## Part 2: Post-Hoc Comparisons

### 2A. Tukey HSD

The ANOVA told us *that* at least one pair of means differs — but not
*which* pair(s). Use Tukey's HSD to identify the specific pairwise
differences.

::: {.callout-tip}
Use `tukey_hsd(rom_improvement ~ protocol)` from the `rstatix` package.
This returns adjusted p-values and confidence intervals for each pairwise
comparison.
:::

```{r}
#| label: tukey-hsd

# Your code here

```

Which pair(s) of protocols show a statistically significant difference
at $\alpha = 0.05$? What is the estimated mean difference and 95% CI
for the significant comparison(s)?

**Answer:**

---

### 2B. Compare adjustment methods

Run the pairwise comparisons using **Bonferroni** correction as well.

::: {.callout-tip}
Use `pairwise_t_test(rom_improvement ~ protocol, p.adjust.method = "bonferroni")`
from `rstatix`.
:::

```{r}
#| label: bonferroni

# Your code here

```

Do the Bonferroni results agree with the Tukey HSD results? Would you
expect them to — and why might they sometimes disagree?

**Answer:**

---

### 2C. Write a complete report

Using your results from Parts 1 and 2, write a short paragraph (3–5
sentences) that a clinical researcher could include in a manuscript.
Include: descriptive statistics for each group, the ANOVA result with
test statistic and p-value, and the specific post-hoc findings.

::: {.callout-tip}
See the "Example write-up" from the ANOVA lecture for a model of what
to include and how to structure it.
:::

**Answer:**

---

## Part 3: Choosing Between Parametric and Nonparametric

### 3A. Explore the pain score variable

The `pain_score_12wk` variable is a self-reported pain rating on a 0–10
scale. Compute summary statistics (n, median, mean, SD) for each protocol
group.

```{r}
#| label: pain-summary

# Your code here

```

---

### 3B. Visualize the distribution

Create boxplots (or histograms) of `pain_score_12wk` by `protocol`.

```{r}
#| label: pain-boxplot

# Your code here

```

Describe the shape of the distributions. Are they approximately
symmetric or clearly skewed?

**Answer:**

---

### 3C. Check ANOVA assumptions for pain scores

Check the equal variance assumption (SD ratio) for pain scores. Also
consider the distributional shape you observed in 3B.

```{r}
#| label: pain-assumptions

# Your code here

```

Based on the distribution shape and the nature of the variable (ordinal
0–10 scale), is ANOVA appropriate for comparing pain scores across
groups? What test should you use instead, and why?

**Answer:**

---

## Part 4: Kruskal-Wallis Test

### 4A. State the hypotheses

State the null and alternative hypotheses for a Kruskal-Wallis test
comparing pain scores across the three rehabilitation protocols.

**Answer:**

$H_0$:

$H_A$:

---

### 4B. Run the Kruskal-Wallis test

Use `kruskal.test()` to test whether median pain scores differ across
the three protocols.

::: {.callout-tip}
The syntax is `kruskal.test(pain_score_12wk ~ protocol, data = knee)`.
Pipe the result to `broom::tidy()` for clean output.
:::

```{r}
#| label: kruskal-wallis

# Your code here

```

Report the test statistic ($H$), degrees of freedom, and p-value.

**Answer:**

---

### 4C. Post-hoc pairwise comparisons

If the Kruskal-Wallis test is significant, follow up with pairwise
Wilcoxon rank-sum tests using an FDR correction.

::: {.callout-tip}
Use `wilcox_test(pain_score_12wk ~ protocol, p.adjust.method = "fdr")`
from `rstatix`.
:::

```{r}
#| label: kw-posthoc

# Your code here

```

Which pair(s) of protocols show significantly different pain scores
after correction?

**Answer:**

---

### 4D. Interpret the results

Write a 2–3 sentence conclusion about pain scores across the
rehabilitation protocols. Report the Kruskal-Wallis result and summarize
the post-hoc findings.

**Answer:**

---

## Part 5: Wilcoxon Signed-Rank Test (Paired Data)

### Background

The Timed Up-and-Go (TUG) test measures how long it takes a patient to
stand up from a chair, walk 3 meters, turn around, walk back, and sit
down. It is a standard functional mobility measure — lower times indicate
better mobility. TUG times are typically **right-skewed** because most
patients cluster near a typical value while a few take considerably longer.

Each patient in this study was measured **before** and **after** their
12-week rehabilitation program, regardless of which protocol they
received. We want to test whether rehabilitation improved TUG times
overall.

---

### 5A. Explore the TUG data

Compute the difference `tug_before - tug_after` for each patient (so that
positive values indicate improvement). Then summarize the differences
with the mean, median, and SD.

```{r}
#| label: tug-explore

# Your code here

```

Also create a histogram or dotplot of the TUG **before** values.
Is the distribution symmetric or skewed?

```{r}
#| label: tug-histogram

# Your code here

```

**Answer:**

---

### 5B. Justify the nonparametric approach

In 2–3 sentences, explain why the Wilcoxon signed-rank test is more
appropriate than a paired *t*-test for these data. Consider the
distribution shape and what you know about TUG timing data.

**Answer:**

---

### 5C. State the hypotheses

State the null and alternative hypotheses for the Wilcoxon signed-rank
test.

**Answer:**

$H_0$:

$H_A$:

---

### 5D. Run the Wilcoxon signed-rank test

Test whether the median TUG time improved after rehabilitation.

::: {.callout-tip}
Create a new variable for the differences, then use
`wilcox_test(tug_diff ~ 1, mu = 0)` from `rstatix`. The `~ 1` tells R
this is a one-sample / paired test against a hypothesized median of 0.
:::

```{r}
#| label: wilcoxon-signed-rank

# Your code here

```

---

### 5E. Interpret the results

Report the test statistic, p-value, and the median improvement in TUG
time.

Write a 2–3 sentence conclusion. Did rehabilitation significantly
improve functional mobility as measured by the TUG test?

**Answer:**

---

## Submission checklist

Before submitting, make sure you have:

- [ ] Data file is in your `data/` folder
- [ ] All code chunks run without errors when rendered
- [ ] Answered all written questions
- [ ] Interpretations are written in full sentences, in context
- [ ] Rendered your HTML successfully
- [ ] Uploaded both `.qmd` and `.html` files
