---
title: "Normal and Poisson Distributions"
subtitle: "BMSC 620 – Distributions of Random Variables (Part 2)"
format:
  revealjs:
    theme: default
    slide-number: true
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Today

- Normal distribution
- Standard normal and Z-scores
- Normal probabilities in R
- Normal approximation to the binomial
- Poisson distribution

## Normal distribution

A random variable X is modeled with a normal distribution if:

- Shape: symmetric, unimodal, bell-shaped
- Center: mean (mu)
- Spread: standard deviation (sigma)

Notation:

X ~ N(mu, sigma)

## Normal density

The probability density function is:

f(x) = (1 / (sigma * sqrt(2*pi))) * exp(-(x - mu)^2 / (2 * sigma^2))

## Visualizing normal distributions

```{r}
x <- seq(-4, 4, length.out = 500)
plot(x, dnorm(x), type = "l", lwd = 2,
     ylab = "Density", xlab = "x",
     main = "Standard normal distribution")
```

## Empirical rule (68–95–99.7)

For a normal distribution:

- ~68% within 1 SD of the mean
- ~95% within 2 SDs
- ~99.7% within 3 SDs

## Standard normal distribution

- Mean = 0
- SD = 1
- Denoted Z

Any normal variable can be converted to Z using:

Z = (X - mu) / sigma

## Z-scores

Interpretation:

- Z = 0: exactly at the mean
- Z = 1: one SD above the mean
- Z = -2: two SDs below the mean

## Normal probabilities in R

For Z ~ N(0, 1):

```{r}
pnorm(1.96)
```

This gives P(Z <= 1.96)

## Examples

```{r}
pnorm(2.67)
1 - pnorm(-0.37)
pnorm(2.46) - pnorm(-2.18)
```

Note: P(Z = 1.53) = 0 for a continuous distribution

## General normal probabilities

Example:

X ~ N(8, 2)

Find P(X > 10)

## Solution

```{r}
1 - pnorm(10, mean = 8, sd = 2)
```

## Normal approximation to the binomial

Recall:

X ~ Binomial(n, p)

When n is large and p is not extreme, the binomial distribution is approximately normal.

## Conditions

Normal approximation is reasonable if:

- n * p >= 10
- n * (1 - p) >= 10

## Mean and SD of binomial

If X ~ Bin(n, p):

- mu = n * p
- sigma = sqrt(n * p * (1 - p))

## Poisson distribution

- Discrete distribution
- Models counts in a fixed interval
- Often used for rare events

Parameter:

- lambda (average rate)

## Poisson probability mass function

For X ~ Poisson(lambda):

P(X = k) = (lambda^k * exp(-lambda)) / k!

## Mean and variance

If X ~ Poisson(lambda):

- Mean = lambda
- Variance = lambda

## Poisson probabilities in R

```{r}
dpois(3, lambda = 5)
ppois(3, lambda = 5)
```

## Example

Suppose deaths occur at an average rate of 5 per year.

Probability of exactly 3 deaths in one year:

```{r}
dpois(3, lambda = 5)
```

## Example: different interval

Probability of 2 deaths in 0.5 years:

```{r}
dpois(2, lambda = 5 * 0.5)
```

## Key takeaways

- Normal distribution models continuous, symmetric data
- Z-scores standardize normal variables
- Normal probabilities are easy in R
- Poisson distribution models counts and rare events
