---
title: "Comparing Means Across Groups"
subtitle: "Analysis of Variance (ANOVA)"
author: "Emile Latour, Nicky Wakim, Meike Niederhausen"
date: "`r library(here); source(here('class_dates.R')); w9d2`"
date-format: long
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v5.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | ANOVA"
    html-math-method: mathjax
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false


knitr::opts_chunk$set(dpi = 300)

library(tidyverse)
library(broom)
library(glue)
library(here)
library(knitr)
library(oibiostat)
library(rstatix)
library(janitor)
library(gt)

library(lamisc)
library(laviz)

data("famuss")

# Set theme for plots
theme_set(laviz::theme_minimal_white(grid = "none",
                                     axis = "xy",
                                     base_size = 14))

set.seed(620)
```

# Learning Objectives

By the end of today's lecture, you will be able to:

1. Explain why we cannot just run many pairwise t-tests when comparing 3 or more group means
2. State the null and alternative hypotheses for a one-way ANOVA
3. Interpret the components of an ANOVA table (SS, MS, F-statistic)
4. Check the assumptions required for ANOVA
5. Run a one-way ANOVA in R and interpret the output
6. Explain the multiple comparisons problem and why post-hoc testing is needed
7. Conduct and interpret Tukey HSD, Bonferroni, and FDR post-hoc tests in R

## Roadmap for Today

::::: columns
::: {.column width="50%"}
**Part 1: Why Not Just Use More t-Tests?**

- Extending our toolkit beyond two groups
- The multiple comparisons problem
- Introducing ANOVA

**Part 2: The Logic of ANOVA**

- The research question and data
- Between-group vs. within-group variability
- The F-statistic and F-distribution
- Reading the ANOVA table
:::

::: {.column width="50%"}
**Part 3: Running ANOVA in R**

- Checking assumptions
- The six-step hypothesis test
- Running `aov()` in R
- Writing a complete conclusion

**Part 4: Post-Hoc Comparisons**

- Which groups are actually different?
- Tukey's HSD
- Bonferroni correction
- False Discovery Rate (FDR)
:::
:::::

# Part 1: Why Not Just Use More t-Tests?

## Where we've been

\

So far our inference toolkit for **numerical outcomes** has been:

\

| Lesson | Test | Groups |
|---|---|:---:|
| 9 | One-sample t-test | 1 |
| 10 | Paired t-test | 2 (matched) |
| 11 | Two-sample t-test | 2 (independent) |

\

**Today's question:** What happens when we want to compare **3 or more** groups?

\

::: {.callout-note icon="false"}
### The motivating question

A genetics study wants to know: does ACTN3 genotype (CC, CT, or TT) affect how much arm strength changes after an exercise intervention?

That's **three groups** — we can't use a two-sample t-test anymore.
:::

## Why not just run three t-tests?

\

With 3 groups (CC, CT, TT), we could run every possible pairwise comparison:

\

::::: columns
::: {.column width="48%"}

**3 tests:**

- CC vs. CT
- CC vs. TT
- CT vs. TT

\

Each test has $\alpha = 0.05$ probability of a Type I error.

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**The problem:** With multiple tests, errors accumulate.

$$P(\text{at least one false positive}) = 1 - (1-\alpha)^m$$

For $m = 3$ tests at $\alpha = 0.05$:

$$1 - (0.95)^3 = 0.143$$

That's a **14% chance** of at least one false positive — not 5%!

:::
:::::

\

<!-- the (1-\alpha)^m formula assumes independent tests, which pairwise t-tests on overlapping groups aren't — so the actual inflation can be even worse -->

::: {.callout-important icon="false"}
### The multiple comparisons problem

Running many tests inflates the Type I error rate beyond the nominal $\alpha$. We need a method that tests all groups simultaneously — that's ANOVA.
:::

## The multiple comparisons problem grows fast

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

m_groups <- 4
n_tests <- choose(4, 2)
p_fp <- 1 - (0.95)^m_groups
p_fp <- lamisc::fmt_pct(p_fp, accuracy = 1.0)

tibble(
  m = 1:20,
  p_error = 1 - (0.95)^m,
  k_groups = round((1 + sqrt(1 + 8*m)) / 2)
) %>%
  ggplot(aes(x = m, y = p_error)) +
  geom_line(linewidth = 1.5, color = "#0072B2") +
  geom_point(size = 3, color = "#0072B2") +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             color = "#E69F00", linewidth = 1) +
  annotate("text", x = 15, y = 0.12,
           label = "Target α = 0.05", color = "#E69F00", size = 5) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 1)) +
  labs(
    title = "P(at least one false positive) inflates with number of tests",
    subtitle = "Each individual test uses α = 0.05",
    x = "Number of pairwise tests (m)",
    y = "P(at least one false positive)"
  )
  
```

**If we had `r m_groups` groups that's `r n_tests` tests → `r p_fp` chance of a false positive. ANOVA solves this by testing all groups in a single test.**

# Part 2: The Logic of ANOVA

## Motivating example: arm strength and genotype

\

::: {.callout-note icon="false"}
### The FAMuSS Study

The **Functional SNPs Associated with Muscle Size and Strength** (FAMuSS) study investigated whether genetic variation is associated with exercise response.

- **Participants:** ~600 young adults completed a resistance training program
- **Genetic variable:** ACTN3 genotype at position R577X (CC, CT, or TT)
- **Outcome:** Percent change in non-dominant arm strength after training

**Question:** Is mean percent change in arm strength different across the three genotype groups?
:::

## Mean arms strength by genotype group

\


```{r}
library(oibiostat)
data("famuss")

famuss %>%
  group_by(actn3.r577x) %>%
  summarise(n = n(),
            mean = round(mean(ndrm.ch), 1),
            sd   = round(sd(ndrm.ch), 1))
```

## Visualizing the three groups

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

ggplot(famuss, aes(x = actn3.r577x, y = ndrm.ch,
                   color = actn3.r577x, fill = actn3.r577x)) +
  geom_boxplot(alpha = 0.1, 
               outlier.shape = NA, 
               width = 0.4) +
  geom_jitter(alpha = 0.5, width = 0.15, size = 1.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 5, color = "black") +
  geom_hline(aes(yintercept = mean(ndrm.ch)),
             linetype = "dashed", linewidth = 0.8,
             color = "grey40") +
  annotate("text", x = 3.7, y = mean(famuss$ndrm.ch) + 8,
           label = "Grand mean", hjust = 1,
           color = "grey40", size = 4.5) +
  scale_color_manual(values = c("#E69F00", "#0072B2", "#009E73"),
                     guide = "none") +
  scale_fill_manual(values = c("#E69F00", "#0072B2", "#009E73"),
                    guide = "none") +
  labs(
    title = "Percent change in non-dominant arm strength by ACTN3 genotype",
    x = "ACTN3 genotype",
    y = "Percent change in arm strength (%)"
  )
```

The diamond ◆ marks the group mean. The dashed line is the grand mean across all participants. Do the groups look different enough to conclude a real effect?

## The key question ANOVA asks

\

Whether or not group means appear different depends on **two things**:

\

::::: columns
::: {.column width="48%"}

**1. How far apart are the group means?**

If the group means are close to the grand mean → not much going on.

If the group means are far from the grand mean → evidence of a real difference.

This is **between-group variability**.

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**2. How much spread is there within each group?**

If observations are tightly clustered around their group mean → easy to tell groups apart.

If observations are widely scattered → hard to tell groups apart even if means differ.

This is **within-group variability**.

:::
:::::

\

::: {.callout-tip icon="false"}
### ANOVA's insight

- ANOVA compares these two sources of variability. 
- If between-group variability is large *relative to* within-group variability, that's evidence that the group means are genuinely different.
:::

## Three sources of variability — visually

\

```{r}
#| echo: false
#| fig-width: 14
#| fig-height: 4.5
#| fig-align: center

grand_mean <- mean(famuss$ndrm.ch)

p_total <- ggplot(famuss, aes(x = actn3.r577x, y = ndrm.ch,
                              color = actn3.r577x)) +
  geom_jitter(alpha = 0.5, width = 0.15, size = 1.5) +
  geom_hline(yintercept = grand_mean, linetype = "dashed",
             linewidth = 1, color = "grey30") +
  scale_color_manual(values = c("#E69F00", "#0072B2", "#009E73"),
                     guide = "none") +
  labs(title = "Total variability (SST)",
       subtitle = "Each point vs. grand mean",
       x = "Genotype", y = "% change") +
  theme(plot.title = element_text(size = 14, face = "bold"))

group_means <- famuss %>%
  group_by(actn3.r577x) %>%
  summarise(grp_mean = mean(ndrm.ch))

famuss_aug <- famuss %>%
  left_join(group_means, by = "actn3.r577x")

p_between <- ggplot(famuss_aug, aes(x = actn3.r577x,
                                    color = actn3.r577x)) +
  geom_jitter(aes(y = ndrm.ch), alpha = 0.1, width = 0.15, size = 1.5) +
  geom_hline(yintercept = grand_mean, linetype = "dashed",
             linewidth = 1, color = "grey30") +
  geom_point(aes(y = grp_mean), shape = 18, size = 6, color = "black") +
  scale_color_manual(values = c("#E69F00", "#0072B2", "#009E73"),
                     guide = "none") +
  labs(title = "Between-group variability (SSG)",
       subtitle = "Group means vs. grand mean",
       x = "Genotype", y = "% change") +
  theme(plot.title = element_text(size = 14, face = "bold"))

p_within <- ggplot(famuss_aug, aes(x = actn3.r577x, y = ndrm.ch,
                                    color = actn3.r577x)) +
  geom_jitter(alpha = 0.5, width = 0.15, size = 1.5) +
  geom_point(aes(y = grp_mean), shape = 18, size = 6, color = "black") +
  scale_color_manual(values = c("#E69F00", "#0072B2", "#009E73"),
                     guide = "none") +
  labs(title = "Within-group variability (SSE)",
       subtitle = "Each point vs. its group mean",
       x = "Genotype", y = "% change") +
  theme(plot.title = element_text(size = 14, face = "bold"))

library(patchwork)
p_total + p_between + p_within
```

$$\underbrace{SST}_{\text{Total}} = \underbrace{SSG}_{\text{Between groups}} + \underbrace{SSE}_{\text{Within groups}}$$

## The ANOVA table

\

ANOVA organizes the variability into a table. $k =$ the number of groups and $N =$ the total number of observations across $k$ groups:

\

::: {style="width: 80%; margin: 0 auto;"}
| Source | SS | df | MS | F |
|---|---|---|---|---|
| Between groups (Model) | SSG | $k-1$ | $MSG = SSG/(k-1)$ | $F = MSG/MSE$ |
| Within groups (Error) | SSE | $N-k$ | $MSE = SSE/(N-k)$ | |
| Total | SST | $N-1$ | | |
:::

\

::: {.callout-note icon="false"}
### Mean Squares: the key step

Dividing each SS by its degrees of freedom converts sums of squares into **mean squares** — averages of squared deviations. This puts between and within variability on the same scale so we can compare them with a ratio.

$$F = \frac{MSG}{MSE} = \frac{\text{between-group variability}}{\text{within-group variability}}$$
:::

## The F-statistic

\

::::: columns
::: {.column width="48%"}

$F = \frac{MSG}{MSE}$

\

**When $H_0$ is true** (all means equal):

- Between-group variability ≈ within-group variability
- $F \approx 1$

**When $H_A$ is true** (at least one mean differs):

- Between-group variability > within-group variability
- $F > 1$, potentially much larger

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5

# ggplot(data.frame(x = c(0, 6)), aes(x = x)) +
#   stat_function(fun = df, args = list(df1 = 2, df2 = 592),
#                 linewidth = 1.5, color = "#0072B2") +
#   geom_vline(xintercept = 3.2, color = "#E69F00",
#              linewidth = 1.2, linetype = "solid") +
#   geom_area(stat = "function", fun = df,
#             args = list(df1 = 2, df2 = 592),
#             fill = "#E69F00", alpha = 0.3,
#             xlim = c(3.2, 6)) +
#   annotate("text", x = 3.5, y = 0.25,
#            label = "F-statistic", hjust = 0,
#            color = "#E69F00", size = 5) +
#   annotate("text", x = 4.2, y = 0.05,
#            label = "p-value", hjust = 0,
#            color = "#E69F00", size = 5) +
#   labs(title = "F-distribution (df1 = 2, df2 = 592)",
#        subtitle = "p-value is always the upper tail",
#        x = "F", y = "Density")


df1 <- length(unique(famuss$actn3.r577x)) - 1
df2 <- dim(famuss)[1] - length(unique(famuss$actn3.r577x))

ggplot(data.frame(x = c(0, 6)), aes(x = x)) +
  stat_function(fun = df, args = list(df1 = df1, df2 = df2),
                linewidth = 1.5, color = "#0072B2") +
  geom_vline(xintercept = 3.2, color = "#E69F00",
             linewidth = 1.2, linetype = "solid") +
  geom_area(stat = "function", fun = df,
            args = list(df1 = 2, df2 = 592),
            fill = "#E69F00", alpha = 0.5,
            xlim = c(3.2, 6)) +
  # Outline the shaded region so it's visible even when thin
  stat_function(fun = df, args = list(df1 = df1, df2 = df2),
                xlim = c(3.2, 6),
                linewidth = 1.2, color = "#E69F00") +
  annotate("text", x = 3.3, y = 0.3,
           label = "F-statistic", hjust = 0,
           color = "#E69F00", size = 5) +
  annotate("text", x = 4.275, y = 0.11,
           label = "p-value", hjust = 0,
           color = "#E69F00", size = 5) +
  annotate("segment", x = 4.25, y = 0.10,
           xend = 3.8, yend = 0.0375,
           curvature = -0.3,
           arrow = arrow(length = unit(0.2, "cm")),
           color = "#E69F00", 
           linewidth = 0.8) +
  labs(title = glue::glue("F-distribution (df1 = {df1}, df2 = {df2})"),
       subtitle = "p-value is always the upper tail",
       x = "F", y = "Density") + 
  scale_y_continuous(expand = c(0, 0))
```



:::
:::::


::: {.callout-important icon="false"}
The p-value for an F-test is **always the upper tail** — we only care about F being surprisingly large.
:::

# Part 3: Running ANOVA in R

## Step 1: State hypotheses

\

::: {.callout-important icon="false"}
### The ANOVA hypotheses

The null hypothesis for ANOVA is that **all** group means are equal:

$$H_0: \mu_{CC} = \mu_{CT} = \mu_{TT}$$

The alternative is that **at least one pair** of means differs:

<!-- $$H_A: \text{At least one pair } \mu_i \neq \mu_j \text{ for } i \neq j$$ -->
$$H_A: \text{At least one mean differs from the others}$$
:::

\

::: {.callout-warning icon="false"}
### Common mistake

People often write $H_A$ as "all means are different." That is **wrong!**

$H_A$ only requires that *at least one* group differs from at least one other. It could be that two groups are identical and a third is different — that still counts as $H_A$.
:::

## Step 1: State hypotheses — in words

\

::::: columns
::: {.column width="48%"}

**In symbols:**

\


$$H_0: \mu_{CC} = \mu_{CT} = \mu_{TT}$$

<!-- $$H_A: \text{At least one pair } \mu_i \neq \mu_j$$ -->
$$H_A: \text{At least one mean differs from the others}$$

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**In words:**

\


$H_0$: The mean percent change in non-dominant arm strength is the same across all three ACTN3 genotype groups.

$H_A$: The mean percent change in non-dominant arm strength differs for at least one pair of ACTN3 genotype groups.

:::
:::::



## Step 2: Set significance level

\

We use the conventional significance level:

$$\alpha = 0.05$$

## Step 3: Check assumptions

\

The F-statistic follows an F-distribution when these conditions hold:

\

**1. Independence:** Observations within and across groups are independent

- Met by design in FAMuSS (different participants in each group)

\

**2. Normality within groups (or large samples):**

- Each group should be approximately normally distributed, OR $n_i \geq 30$ per group

```{r}
famuss %>%
  group_by(actn3.r577x) %>%
  summarise(n = n())
```

All groups have $n > 30$ ✓

## Step 3: Check assumptions (continued)

\

**3. Equal variance across groups (homoscedasticity):**

- Within-group variability should be similar across groups
- Rule of thumb: violated if $\max(s_i) / \min(s_i) > 2$

```{r}
famuss %>%
  group_by(actn3.r577x) %>%
  summarise(sd = sd(ndrm.ch)) %>%
  summarise(ratio = max(sd) / min(sd))
```

\

Ratio is close to 1 — equal variance assumption is reasonable ✓


::: {.callout-tip icon="false"}
If the equal variance assumption is seriously violated, Welch's ANOVA (`oneway.test(..., var.equal = FALSE)`) is a robust alternative that does not require equal variances.
:::

```{r}
#| echo: false

# ggplot(data = famuss, 
#        aes(x = ndrm.ch)) + 
#   geom_histogram(binwidth = 10, 
#                  colour = "white") + 
#   stat_function(fun = dnorm) + 
#   facet_wrap(~ actn3.r577x)
# 
# # Pre-compute normal curves for each group
# norm_curves <- famuss %>%
#   group_by(actn3.r577x) %>%
#   summarise(mean = mean(ndrm.ch), sd = sd(ndrm.ch)) %>%
#   rowwise() %>%
#   mutate(x = list(seq(-50, 150, length.out = 200))) %>%
#   unnest(x) %>%
#   mutate(y = dnorm(x, mean, sd))
# 
# ggplot(famuss, aes(x = ndrm.ch)) +
#   geom_histogram(aes(y = after_stat(density)),
#                  binwidth = 10, colour = "white") +
#   geom_line(data = norm_curves, aes(x = x, y = y),
#             color = "#E69F00", linewidth = 1.2) +
#   facet_wrap(~ actn3.r577x)
```


## Steps 4–5: Calculate F-statistic and p-value

\

```{r}
# Fit the ANOVA model
famuss_aov <- aov(ndrm.ch ~ actn3.r577x, data = famuss)

# View the ANOVA table
summary(famuss_aov)
```

\

::: {.callout-note icon="false"}
### Reading the output

- `actn3.r577x` row = **between-group** (model) variability
- `Residuals` row = **within-group** (error) variability
- `F value` = MSG / MSE
- `Pr(>F)` = p-value (upper tail of F-distribution)
:::

## Steps 4–5: Tidy output with `broom`

\

For inline reporting or downstream use, `tidy()` from the `broom` package converts the ANOVA table to a data frame:

```{r}
famuss_aov %>%
  tidy()
```

\

Pull the p-value for inline reporting:

```{r}
famuss_aov %>%
  tidy() %>%
  filter(term == "actn3.r577x") %>%
  pull(p.value) %>%    # From dplyr package. Extract ("pull") a single column 
  round(4)
```





## Step 6: Conclusion

\

```{r}
#| echo: false

p_val <- famuss_aov %>%
  tidy() %>%
  filter(term == "actn3.r577x") %>%
  pull(p.value) %>%
  scales::pvalue()


f_stat <- famuss_aov %>%
  tidy() %>%
  filter(term == "actn3.r577x") %>%
  pull(statistic) %>%
  scales::number(accuracy = 0.01)


dfs <- famuss_aov %>%
  tidy() %>%
  pull(df)


mean_sd <- famuss %>% 
  group_by(actn3.r577x) %>% 
  lamisc::calc_summ_stats(ndrm.ch) %>% 
  mutate(mean = lamisc::fmt_pct(mean, accuracy = 0.1, scale = 1)) %>% 
  mutate(sd = lamisc::fmt_pct(sd, accuracy = 0.1, scale = 1)) %>% 
  mutate(mean_sd = glue::glue("{mean} (SD = {sd})")) %>% 
  dplyr::pull(mean_sd)
```

Recall our hypotheses:

$$H_0: \mu_{CC} = \mu_{CT} = \mu_{TT} \qquad \text{vs.} \qquad H_A: \text{At least one mean differs from the others}$$

\

With $F$ = `r f_stat`, $df$ = (`r dfs[1]`, `r dfs[2]`), *p* = `r p_val`, and $\alpha = 0.05$: We **reject $H_0$**.

\

::: {.callout-important icon="false"}
### Conclusion

There is sufficient evidence that the mean percent change in non-dominant arm strength differs for at least one pair of ACTN3 genotype groups ($F$ = `r f_stat`, $df$ = `r dfs[1]`, `r dfs[2]`, *p* = `r p_val`).

Descriptively, mean arm strength change was `r mean_sd[1]` for CC, `r mean_sd[2]` for CT, and `r mean_sd[3]` for TT.
:::

## What ANOVA tells us — and what it doesn't (1/2)

\

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.9
#| fig-align: center

famuss %>%
  group_by(actn3.r577x) %>%
  summarise(mean = mean(ndrm.ch), sd = sd(ndrm.ch), n = n()) %>%
  mutate(se = sd / sqrt(n),
         lo = mean - 1.96 * se,
         hi = mean + 1.96 * se) %>%
  ggplot(aes(x = actn3.r577x, y = mean, color = actn3.r577x)) +
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.075, linewidth = 1.2) +
  scale_color_manual(values = c("#E69F00", "#0072B2", "#009E73"),
                     guide = "none") +
  labs(title = "Group means with 95% CIs",
       subtitle = "Significant ANOVA — but which pair(s) differ?",
       x = "ACTN3 genotype", y = "Mean % change in arm strength")
```

## What ANOVA tells us — and what it doesn't (2/2)

\

::: {.callout-warning icon="false"}
### A critical limitation

ANOVA tells us that **at least one pair** of means differs. It does **not** tell us:

- Which specific groups are different from which
- How many pairs differ
- The direction or magnitude of the differences

After a significant ANOVA, we need **post-hoc tests** to identify where the differences lie.
:::




# Part 4: Post-Hoc Comparisons

## The multiple comparisons problem, revisited

\

After a significant ANOVA, we want to test all pairwise comparisons. With $k = 3$ groups:

$${3 \choose 2} = 3 \text{ pairwise tests: CC vs. CT, CC vs. TT, CT vs. TT}$$

\

If we just ran three regular t-tests at $\alpha = 0.05$:

$$P(\text{at least one false positive}) = 1 - (0.95)^3 = 0.143$$

\

**Post-hoc methods control this inflated error rate** through different strategies — each with different tradeoffs between Type I and Type II error.

## Post-hoc comparisons with `rstatix`

\

The `rstatix` package provides tidy post-hoc testing:

\

```{r}

# Tukey HSD — controls family-wise error rate
famuss %>%
  tukey_hsd(ndrm.ch ~ actn3.r577x)
```

## Tukey's Honest Significant Difference (HSD)

\

::::: columns
::: {.column width="55%"}

**What it does:**

- Controls the **family-wise error rate** (FWER) — the probability of making *any* false positive across all comparisons
- Compares all pairwise differences simultaneously
- Reports both adjusted p-values and confidence intervals for each difference
- **The standard choice for ANOVA post-hoc testing**

:::

::: {.column width="5%"}
:::

::: {.column width="40%"}

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

# famuss_aov %>%
#   TukeyHSD() %>%
#   tidy() %>%
#   ggplot(aes(x = contrast,
#              y = estimate,
#              ymin = conf.low,
#              ymax = conf.high,
#              color = adj.p.value < 0.05)) +
#   geom_pointrange(linewidth = 1.2, size = 0.8) +
#   geom_hline(yintercept = 0, linetype = "dashed",
#              color = "grey40") +
#   scale_color_manual(values = c("grey60", "#0072B2"),
#                      labels = c("Not significant", "Significant"),
#                      name = NULL) +
#   coord_flip() +
#   labs(title = "Tukey HSD",
#        subtitle = "95% family-wise CIs",
#        x = NULL, y = "Difference in means")

famuss_aov %>%
  rstatix::tukey_hsd() %>%
  mutate(contrast = glue::glue("{group1}-{group2}")) %>%
  ggplot(aes(x = contrast,
             y = estimate,
             ymin = conf.low,
             ymax = conf.high,
             color = p.adj < 0.05)) +
  geom_pointrange(linewidth = 1.2, size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey40") +
  scale_color_manual(values = c("grey60", "#0072B2"),
                     labels = c("Not significant", "Significant"),
                     name = NULL) +
  coord_flip() +
  labs(title = "Tukey HSD",
       subtitle = "95% family-wise CIs",
       x = NULL, y = "Difference in means")
```

:::
:::::

Which comparison is driving the significant ANOVA result?

## Bonferroni correction


**Strategy:** Divide $\alpha$ by the number of tests: $\alpha_{Bonf} = \alpha / m$

Equivalently, multiply each p-value by $m$ and compare to $\alpha$.

```{r}
# Bonferroni correction
famuss %>%
  rstatix::pairwise_t_test(ndrm.ch ~ actn3.r577x,
                           p.adjust.method = "bonferroni")
```

\

::: {.callout-note icon="false"}
### Bonferroni: pros and cons

**Pro:** Simple, transparent, very conservative (strong Type I error control)

**Con:** Loses power quickly as the number of comparisons grows — becomes hard to detect real differences

Bonferroni is best when you have a small number of *a priori* comparisons you care about.
:::

## False Discovery Rate (FDR)

\

**Strategy:** Control the *expected proportion* of false positives among significant results, rather than the probability of *any* false positive.

```{r}
# Benjamini-Hochberg FDR
famuss %>%
  rstatix::pairwise_t_test(ndrm.ch ~ actn3.r577x,
                           p.adjust.method = "fdr")
```

\

::: {.callout-note icon="false"}
### FDR: pros and cons

**Pro:** More powerful than Bonferroni or Tukey when there are many comparisons — allows more discoveries

**Con:** Some false positives are tolerated (you control the *rate*, not the *occurrence*)

FDR is especially popular in genomics and omics where hundreds or thousands of tests are run simultaneously.
:::

## Comparing the three approaches

\

| Method | Controls | Best used when | Conservativeness |
|---|---|---|:---:|
| **Tukey HSD** | Family-wise error rate | Post-ANOVA pairwise comparisons | Moderate |
| **Bonferroni** | Family-wise error rate | Few *a priori* comparisons | High |
| **FDR** | False discovery rate | Many simultaneous tests | Low |

\

::: {.callout-tip icon="false"}
### Rule of thumb for BMSC 620

For ANOVA post-hoc testing in a standard biomedical study: **Tukey HSD** is your default.

Use Bonferroni when you have a small number of planned comparisons stated *before* looking at the data.

Use FDR when you're doing many tests (e.g., comparing gene expression across dozens of groups).
:::

## Visualizing post-hoc results (1/2)

```{r}
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
#| eval: false

# Add significance brackets to a boxplot using rstatix + ggpubr
library(ggpubr)

tukey_results <- famuss %>%
  rstatix::tukey_hsd(ndrm.ch ~ actn3.r577x) %>%
  rstatix::add_xy_position(x = "actn3.r577x")

ggpubr::ggboxplot(famuss, x = "actn3.r577x", y = "ndrm.ch",
                  color = "actn3.r577x",
                  width = 0.5, 
                  palette = c("#E69F00", "#0072B2", "#009E73"),
                  add = "jitter", add.params = list(alpha = 0.3)) +
  ggpubr::stat_pvalue_manual(tukey_results, hide.ns = TRUE) +
  labs(title = "Percent change in arm strength by ACTN3 genotype",
       subtitle = "Tukey HSD post-hoc comparisons (significant pairs shown)",
       x = "ACTN3 genotype",
       y = "Percent change in arm strength (%)",
       color = "Genotype")




```

## Visualizing post-hoc results (2/2)

```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center
#| echo: false

# Add significance brackets to a boxplot using rstatix + ggpubr
library(ggpubr)

tukey_results <- famuss %>%
  rstatix::tukey_hsd(ndrm.ch ~ actn3.r577x) %>%
  rstatix::add_xy_position(x = "actn3.r577x")

ggpubr::ggboxplot(famuss, x = "actn3.r577x", y = "ndrm.ch",
                  color = "actn3.r577x",
                  width = 0.5, 
                  palette = c("#E69F00", "#0072B2", "#009E73"),
                  add = "jitter", add.params = list(alpha = 0.3)) +
  ggpubr::stat_pvalue_manual(tukey_results, hide.ns = TRUE) +
  labs(title = "Percent change in arm strength by ACTN3 genotype",
       subtitle = "Tukey HSD post-hoc comparisons (significant pairs shown)",
       x = "ACTN3 genotype",
       y = "Percent change in arm strength (%)",
       color = "Genotype")




```

## Putting it all together: full reporting

\

A complete report of this analysis would include:

\

::: {.callout-important icon="false"}
### Example write-up

```{r}
#| echo: false

tukey_res <- famuss %>%
  tukey_hsd(ndrm.ch ~ actn3.r577x)

tukey_res <- tukey_res %>% 
  mutate(dplyr::across(.cols = c(estimate, 
                                 conf.low, 
                                 conf.high), 
                       .fns = ~ lamisc::fmt_num(x = ., 
                                                accuracy = 0.1)),
        p.adj = lamisc::fmt_pvl(p.adj))


# Filter to the comparison of interest
tt_vs_cc <- tukey_res %>%
  filter(group1 == "CC", group2 == "TT")

```


Mean percent change in non-dominant arm strength was `r mean_sd[1]` for CC, `r mean_sd[2]` for CT, and `r mean_sd[3]` for TT genotype groups. A one-way ANOVA indicated a statistically significant difference in mean arm strength change across genotype groups ($F$(`r dfs[1]`, `r dfs[2]`) = `r f_stat`, *p* = `r p_val`).

Post-hoc comparisons using Tukey's HSD revealed that the TT group had significantly greater mean arm strength change than the CC group (difference = `r tt_vs_cc$estimate` percentage points, 95% CI: `r tt_vs_cc$conf.low` to `r tt_vs_cc$conf.high`, adjusted *p* = `r tt_vs_cc$p.adj`). No significant differences were found between the CC and CT groups (adjusted *p* = `r purrr::pluck(tukey_res, "p.adj", 1)`) or the CT and TT groups (adjusted p = `r purrr::pluck(tukey_res, "p.adj", 3)`).
:::

\

Note what's included: group-level descriptive statistics, the F-statistic with both df values, the p-value, and specific post-hoc results with effect sizes and CIs.

# Wrap-Up

## ANOVA summary

::::: columns
::: {.column width="48%"}

**When to use ANOVA:**

- Continuous outcome
- 3 or more independent groups
- Groups are independent of each other

\

**The hypotheses:**

$$H_0: \mu_1 = \mu_2 = \cdots = \mu_k$$
<!-- $$H_A: \text{at least one pair } \mu_i \neq \mu_j$$ -->
$$H_A: \text{At least one mean differs from the others}$$


\

**Key assumption to check:**

- Similar within-group SDs ($SD_{max} / SD_{min}$ ratio < 2)
- $n \geq 30$ per group or approximate normality

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

**R workflow:**

```{r}
#| eval: false

# 1. Check assumptions
famuss %>%
  group_by(actn3.r577x) %>%
  summarise(n = n(), sd = sd(ndrm.ch))

# 2. Fit ANOVA
famuss_aov <- aov(ndrm.ch ~ actn3.r577x,
                  data = famuss)
summary(famuss_aov)

# 3. Post-hoc (if significant)
famuss %>%
  tukey_hsd(ndrm.ch ~ actn3.r577x)
```


**Post-hoc methods:**

| Method | Use when |
|---|---|
| `tukey_hsd()` | Default for ANOVA |
| `rstatix::pairwise_t_test` |  |
| `p.adjust = "bonferroni"` | Few planned comparisons |
| `p.adjust = "fdr"` | Many tests |

:::
:::::

## Connecting to what's next

\

**Lesson 17 (Monday):** Nonparametric tests

- What if the normality assumption is violated *and* sample sizes are small?
- Mann-Whitney U, Kruskal-Wallis — the nonparametric analogues to the t-test and ANOVA

\

**Lesson 18 (Wednesday):** Correlation and Simple Linear Regression

- What if both variables are continuous?
- Extending beyond group comparisons to relationships between variables
- ANOVA is actually a special case of regression — we'll see the connection

\

::: {.callout-tip icon="false"}
### The big picture

ANOVA, t-tests, and linear regression are all part of the same family. The F-test in ANOVA and the t-test are mathematically related — in fact, for $k = 2$ groups, $F = t^2$. 
:::
