---
title: "Odds & Ends: Statistics for Animal Studies"
subtitle: "What you know, what to watch out for, and where to go next"
author: "Emile Latour"
date: "`r library(here); source(here('class_dates.R')); w11d2`"
date-format: long
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v5.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | Statistics for Animal Studies"
    html-math-method: mathjax
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(broom)
library(glue)
library(here)
library(knitr)
library(gt)
library(pwr)
library(lamisc)
library(laviz)

theme_set(laviz::theme_minimal_white(grid = "none",
                                     axis = "xy",
                                     base_size = 16))

# Simulated tumor growth data for the longitudinal example
set.seed(620)
n_mice <- 6

tumor_data <- tibble(
  mouse_id  = rep(paste0("M", 1:n_mice), each = 4),
  treatment = rep(c("Control", "Control", "Control",
                    "Treated", "Treated", "Treated"), each = 4),
  week      = rep(c(1, 2, 3, 4), times = n_mice),
  volume    = c(
    # Control mice — growing fast
    120 + cumsum(rnorm(4, 30, 8)),
    110 + cumsum(rnorm(4, 28, 7)),
    130 + cumsum(rnorm(4, 32, 9)),
    # Treated mice — slower growth, different baseline
    100 + cumsum(rnorm(4, 10, 6)),
    115 + cumsum(rnorm(4, 12, 7)),
     95 + cumsum(rnorm(4, 11, 5))
  )
)

# Naive: t-test at each timepoint
week4_data <- tumor_data %>% filter(week == 4)

# Power calculation values for 3Rs slide
pwr_result <- pwr.t.test(d = 0.8, sig.level = 0.05, power = 0.80,
                          type = "two.sample")
n_per_group <- ceiling(pwr_result$n)
```

# Part 1: You Know More Than You Think

## Mapping your experiments to your toolkit

\

Every animal study comes down to a few fundamental questions. You've been answering versions of these all semester.

\

::::: columns
::: {.column width="60%"}

The key decisions are ones you already know how to make:

\

::: {.callout-note icon="false"}
### The three questions that determine your test
1. **What type of outcome?** Continuous, binary, ordinal?
2. **How many groups?** One, two, or three or more?
3. **Are the observations independent?** Or are they paired / repeated on the same subject?
:::

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center

tibble(
  question = c("Continuous\noutcome?", "Two\ngroups?", "Paired\nobservations?"),
  x = c(1, 2, 3),
  y = c(1, 1, 1)
) %>%
  ggplot(aes(x = x, y = y, label = question)) +
  geom_label(size = 5, fill = "#0072B2", color = "white",
             fontface = "bold", label.size = 0,
             label.padding = unit(0.6, "lines")) +
  geom_segment(aes(x = 1.35, xend = 1.675, y = 1, yend = 1),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1,
               color = "#555555") +
  geom_segment(aes(x = 2.275, xend = 2.55, y = 1, yend = 1),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1,
               color = "#555555") +
  xlim(0.5, 3.5) + ylim(0.5, 1.5) +
  theme_void()
```

:::
:::::

## Your course toolkit, translated to the bench

\

::: {style="font-size: 0.82em;"}

| Study design | Groups | Outcome | Test | Key R function |
|---|:---:|:---:|---|---|
| Compare mean to a known reference value | 1 | Continuous, ~normal | One-sample t-test | `t.test(y, mu = ref)` |
| Two groups, one timepoint (~normal) | 2 | Continuous, ~normal | Two-sample t-test | `t.test(y ~ group)` |
| Two groups, one timepoint, skewed or small $n$ | 2 | Continuous, skewed | Wilcoxon rank-sum | `wilcox.test(y ~ group)` |
| Same animal before and after treatment | 1 (paired) | Continuous | Paired t-test | `t.test(pre, post, paired = TRUE)` |
| Three or more dose groups at one timepoint | 3+ | Continuous, ~normal | One-way ANOVA | `aov()` + post-hoc |
| Three or more groups, skewed or small $n$ | 3+ | Continuous, skewed | Kruskal-Wallis | `kruskal.test()` |
| Does response rate match a historical benchmark? | 1 | Binary | One-proportion z-test | `prop.test(x, n, p = ref)` |
| Does response rate differ between two groups? (large $n$) | 2 | Binary | Two-proportion z-test | `prop.test()` |
| Does response rate differ between two groups? (small $n$) | 2 | Binary | Fisher's exact | `fisher.test()` |
| Same animal, binary outcome before and after (paired binary) | 1 (paired) | Binary | McNemar's test | `mcnemar.test()` |
| Association between two categorical variables | 2+ | Categorical | Chi-square | `chisq.test()` |
| Continuous predictor and continuous outcome | — | Continuous | Simple linear regression | `lm()` + `tidy()` |

: {.striped tbl-colwidths="[30, 9, 16, 18, 17]"}

:::

## The conversation this prepares you for

\

When you sit down with a statistician, the most useful thing you can do is walk in with clear answers to:

\

::::: columns
::: {.column width="55%"}

::: {.callout-note icon="false"}
### What a statistician wants to know
- What is your **outcome**, and how is it measured?
- What is your **primary comparison** — treatment vs. control? Before vs. after?
- How many **groups**, and are they **independent or paired**?
- What is your **sample size**, and how was it chosen?
- Are there **other variables** that might confound the relationship?
:::

:::

::: {.column width="45%"}

::: {.callout-important icon="false"}

### What this course gave you

You now have the vocabulary to answer all of these questions — and to understand the statistician's answer.

\

**You don't need to know the method. You need to know your data and your question.**

That's what brings to a productive collaboration.
:::

:::
:::::

# Part 2: Experimental Design Matters

## The decision that can't be fixed in analysis

\

::::: columns
::: {.column width="55%"}

Imagine you're preparing slides from tumor tissue.

\

It's a long protocol. You decide to process all **control samples on Monday** and all **treated samples on Wednesday**.

\

Seems fine — same technician, same reagents, same protocol.

\

**But what if:**

- The antibody lot was slightly different?
- Room temperature differed?
- You were more fatigued by slide 30?

:::

::: {.column width="45%"}

::: {.callout-important icon="false"}
### The problem: confounding by batch

Now **treatment** and **processing day** are perfectly confounded.

Any difference you observe between groups could be due to:

- The treatment ✓
- The day of processing ✗
- Some combination of both ✗

**You cannot separate them in analysis.** The damage is done at the design stage.
:::

:::
:::::

## What is a batch effect?

\

A **batch effect** is any source of technical variability introduced by when or how samples are processed — not by your biological treatment.


::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Common sources of batch effects
- Day of experiment / run
- Reagent lot or antibody batch
- Technician (inter-operator variability)
- Equipment calibration drift
- Passage number in cell lines
- Time in culture before harvest
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Why cell lines deserve special mention
Cell lines can drift over passages. A control run at passage 12 and a treatment run at passage 20 may not be biologically comparable — even if the protocol is identical.

This is a batch effect with biological consequences.
:::

:::
:::::

\

::: {.callout-tip icon="false"}
### The fix: blocking
**Randomize** your treatment and control samples across batches — process some controls and some treated samples together in each run. Now batch effects cancel out across groups rather than accumulating in one of them.
:::

## Blocking: turning nuisance into noise

\

::::: columns
::: {.column width="55%"}

**Without blocking:** treatment is confounded with day

```
Monday:    Control  Control  Control
Wednesday: Treated  Treated  Treated
```

You can't know if differences come from treatment or day.

\

**With blocking:** treatment is balanced across days

```
Monday:    Control  Treated  Control  Treated
Wednesday: Treated  Control  Treated  Control
```

Now any day effect hits both groups equally — it's **noise**, not **bias**.

\

::: {.callout-note icon="false"}
Blocking is the experimental design analog of the **paired t-test** — you're using structure in the design to reduce unwanted variability.
:::

:::

::: {.column width="45%"}

::: {.callout-important icon="false"}
### The principle

**Design your experiment so that any batch effect is balanced across your groups.**

\

This is why randomizing *within* batches — not just randomizing overall — matters. Convenience-driven sequencing (all controls first, then all treated) is one of the most common and consequential design errors in bench science.
:::

:::
:::::

# Part 3: The Replication Problem

## Not all replicates are created equal

\

::::: columns
::: {.column width="50%"}

::: {.callout-note icon="false"}
### Biological replicates
**Separate biological units** — different animals, different cell culture preparations, different patients.

Each is an independent sample from the population of interest.

**This is what your sample size $n$ should count.**
:::

\

::: {.callout-note icon="false"}
### Technical replicates
**Repeated measurements of the same biological unit** — running the same sample through the assay three times, imaging the same tissue section twice.

These measure **instrument precision**, not biological variability.
:::

:::

::: {.column width="50%"}

::: {.callout-important icon="false"}
### Why it matters statistically

If you treat technical replicates as biological replicates, you **artificially inflate your sample size**.

\

**Example:** 3 mice, each measured 5 times.

- True biological $n$ = **3**
- If you treat all measurements as independent: $n$ = **15**

With $n = 15$ your standard errors are artificially small, your t-statistics are inflated, and your p-values are too small.

You will find "significant" results that aren't real.
:::

:::
:::::

## What the field actually does — and why

\

::::: columns
::: {.column width="55%"}

::: {.callout-tip icon="false"}
### An honest conversation

Using technical replicates as if they were biological replicates is **not statistically correct**.

It is, however, **common in practice** — particularly in early-stage, exploratory bench work.

\

Why? Because biological replicates are expensive, time-consuming, and sometimes require institutional ethics approval for additional animals.

\

The pragmatic approach many labs take: average technical replicates within each biological unit first, then analyze those averages. This at least preserves the correct $n$.
:::

:::

::: {.column width="45%"}

::: {.callout-important icon="false"}
### What you should do

1. **Be clear** about what type of replicates you have
2. **Report** both biological and technical $n$ in your methods
3. **Analyze** at the level of the biological replicate
4. **Acknowledge** limitations if technical replicates were used due to feasibility constraints

When in doubt — **ask a statistician before you collect data**, not after.
:::

:::
:::::

# Part 4: Where Your Toolkit Runs Out

## Longitudinal animal studies: the naive approach fails

\

Many animal studies measure the **same outcome repeatedly over time** — tumor volume weekly, body weight daily, behavioral scores at multiple timepoints.


::: {.callout-note icon="false"}
### The tempting but wrong approach
Run a t-test at **each timepoint separately**.
:::

\

::::: columns
::: {.column width="50%"}

**Two problems:**

1. **Multiple comparisons** — testing at 4 timepoints with $\alpha = 0.05$ means your family-wise false positive rate is already $> 18\%$ under the null, just like the ANOVA post-hoc problem
2. **Non-independence** — measurements from the same mouse at week 1 and week 2 are **correlated**. A mouse with high tumor volume at week 1 will likely have high tumor volume at week 2. Standard errors that assume independence are **too small**.

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center

tumor_data %>%
  ggplot(aes(x = week, y = volume,
             color = treatment, group = mouse_id)) +
  geom_line(alpha = 0.6, linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Control" = "#E69F00",
                                "Treated" = "#0072B2")) +
  labs(title = "Tumor volume over time",
       subtitle = "Each line = one mouse",
       x = "Week", y = "Tumor volume (mm³)",
       color = NULL) +
  theme(legend.position = "top")
```

:::
:::::

## The solution: linear mixed models

\

**Linear mixed models** (LMM) handle both problems at once. The key idea: give each animal **its own baseline**.

\

::::: columns
::: {.column width="50%"}

The model has two kinds of effects:

::: {.callout-note icon="false"}
### Fixed effects
The parameters we care about — treatment group, time, their interaction. These are estimated the same way as in regular regression.
:::

::: {.callout-note icon="false"}
### Random effects
Animal-specific offsets from the overall intercept. Each mouse gets its own "starting point," which absorbs the within-animal correlation.
:::

:::

::: {.column width="50%"}

::: {style="font-size: 0.75em;"}

```r
library(lme4)

model_lmm <- lmer(
  volume ~ week * treatment + (1 | mouse_id),
  data = tumor_data
)

summary(model_lmm)
```
:::


\

::: {.callout-tip icon="false"}
The `(1 | mouse_id)` term is the random intercept — it tells R that observations from the same mouse are correlated.

You don't need to know how to fit this yet. You need to know **when to ask for it**.
:::

:::
:::::

## When to ask for a mixed model

\

::: {.callout-important icon="false"}
### The signal that you need more than a t-test or ANOVA

**If the same subject appears more than once in your dataset, you probably need a mixed model.**
:::

\

| Design | Repeated measures? | Method |
|---|:---:|---|
| Two groups, one sacrifice timepoint | No | t-test / Wilcoxon |
| Pre/post on same animal (two timepoints) | Yes (2 only) | Paired t-test |
| Same animal measured weekly for 8 weeks | Yes (many) | Linear mixed model |
| Cells from same animal run in triplicate | Yes (technical) | Mixed model preferred; averaging within animal then t-test is a common simplification |

: {tbl-colwidths="[35, 25, 35]"}


::: {.callout-tip icon="false"}
### What to tell a statistician
*"I have repeated measurements on the same animals over time. I think I need a mixed model but I'm not sure how to set it up."*

That sentence will make any statistician very happy to work with you.
:::

# Part 5: Power and the 3Rs

## Why sample size matters more in animal studies

\

::::: columns
::: {.column width="55%"}

You've used `pwr` to calculate sample size for t-tests and ANOVA. The same tools apply to animal studies — but the **stakes are different**.

\

In a clinical trial, enrolling more patients than needed is a resource question.

In an animal study, it's an **ethical question**.

\

::: {.callout-note icon="false"}
### The 3Rs framework (Russell & Burch, 1959)
- **Replace** — use non-animal alternatives where possible
- **Reduce** — use the minimum number of animals needed to achieve scientific objectives
- **Refine** — minimize suffering and improve animal welfare

**Institutional animal ethics review** and many journals now require a power justification for $n$.
:::

:::

::: {.column width="45%"}

::: {.callout-important icon="false"}
### The two failure modes

**Too few animals (underpowered)**
Study misses a real effect. Animals were used and the science learned nothing. This violates *Reduce*.

\

**Too many animals (overpowered)**
More animals than needed to answer the question. Also violates *Reduce*.

\

Power analysis is not just a statistics formality — **it is an ethical obligation.**
:::

:::
:::::

## Power analysis with tools you already have

\

You already know `pwr.t.test()`. The inputs are the same for animal studies:

\

::::: columns
::: {.column width="55%"}

::: {style="font-size: 0.85em;"}

```{r}
library(pwr)

# How many mice per group do I need?
# Expecting a large effect (d = 0.8),
# alpha = 0.05, power = 0.80

pwr.t.test(
  d         = 0.8,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample"
)
```

:::

:::

::: {.column width="45%"}

::: {.callout-note icon="false"}
### Result
**`r n_per_group` mice per group** — about `r n_per_group * 2` total.

This is the minimum to detect a large effect with 80% power. For a medium effect ($d = 0.5$), you would need substantially more.
:::

:::
:::::

## The hardest part: choosing your effect size

\

> *"I don't know what effect size to expect. What should I use?"*

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Practical approaches
- **Pilot data** from your own lab
- **Published literature** in similar models
- **Smallest effect worth detecting** — what difference would be biologically or clinically meaningful?
- **Cohen's conventions** ($d$ = 0.2 / 0.5 / 0.8) as a last resort — but justify why
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-important icon="false"}
### What not to do
- Don't run the experiment first and then compute power from your results — **post-hoc power is circular**
- Don't choose $n$ by convention ("we always do 5 per group") without justification

\

**Pilot experiments are valuable precisely because they give you an estimate of effect size and variability before you commit to a full study.**
:::

:::
:::::



## Putting it all together

\

::::: columns
::: {.column width="55%"}

**Before you run your experiment:**

1. Define your outcome and primary comparison clearly
2. Identify your design: independent groups? paired? longitudinal?
3. Think about batch effects — plan to randomize within batches
4. Distinguish biological from technical replicates — know your true $n$
5. Run a power analysis — justify your $n$ for ethics review and grant applications

**When the data are in:**

6. Apply the test that matches your design (use the table!)
7. If you have repeated measures — ask for help with mixed models
8. Report biological $n$, test statistic, p-value, and effect size

:::

::: {.column width="45%"}

::: {.callout-tip icon="false"}
### The most valuable thing this class gave you

You can now walk into a meeting with a statistician and have a real conversation.

You know what your outcome is. You know whether it's paired. You know what a p-value means and what it doesn't. You know that sample size requires justification.

**That puts you ahead of most people who show up with their data already collected and a question that can no longer be answered.**
:::

:::
:::::

## On the horizon: two designs worth knowing exist

\

Two extensions of ANOVA come up often in experimental biology. You don't need to know
how to run them today — but you should know when to look them up.

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Two-way ANOVA
**When you need it:** Two categorical explanatory variables, one continuous outcome.

**Example:** Does tumor volume differ by *treatment* AND by *sex*? And does the
treatment effect differ between sexes? That last question is an **interaction**.

**R function:** `aov(y ~ A * B, data = data)`
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Repeated measures ANOVA
**When you need it:** Same subjects at 3+ timepoints; simpler alternative to a
full mixed model when data are complete and balanced.

**Example:** Body weight at baseline, week 2, week 4, and week 8 in the same animals.

**R function:** `aov(y ~ time + Error(id/time), data = data)`
:::

:::
:::::

## Where to learn more

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Free online textbook
**Zhang & Wang (2017), *Advanced Statistics Using R*, Ch. 7**

Clear, R-based, covers one-way, two-way, and repeated measures ANOVA.

[advstats.psychstat.org/book/ZhangWang2017.pdf](https://advstats.psychstat.org/book/ZhangWang2017.pdf)
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Penn State open course notes
**STAT 502** — Analysis of Variance and Design of Experiments

[online.stat.psu.edu/stat502](https://online.stat.psu.edu/stat502)

**STAT 503** — Design of Experiments

[online.stat.psu.edu/stat503](https://online.stat.psu.edu/stat503)
:::

:::
:::::

\

