---
title: "Odds & Ends: Multiple Linear Regression"
subtitle: "Categorical Predictors + Adjusting for Confounding"
author: "Emile Latour"
date: "`r library(here); source(here('class_dates.R')); w11d2`"
date-format: long
format:
  revealjs:
    theme: "../../assets/css/reveal-bmsc620_v5.scss"
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "BMSC 620 | Multiple Linear Regression"
    html-math-method: mathjax
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(broom)
library(glue)
library(here)
library(knitr)
library(oibiostat)
library(rstatix)
library(janitor)
library(gt)
library(lamisc)
library(laviz)

theme_set(laviz::theme_minimal_white(grid = "none",
                                     axis = "xy",
                                     base_size = 16))

data(nhanes.samp)

nhanes_bp <- nhanes.samp %>%
  filter(!is.na(BMI), !is.na(BPSys1), !is.na(Gender), !is.na(Age)) %>%
  select(BMI, BPSys1, Gender, Age) %>%
  mutate(
    Gender   = factor(Gender),   # "female" is reference (alphabetical)
    Age_group = cut(Age,
                    breaks = c(0, 39, 59, Inf),
                    labels = c("Under 40", "40–59", "60+"))
  )

# SLR model (for comparison)
model_slr <- lm(BPSys1 ~ BMI, data = nhanes_bp)

# Categorical model
model_cat <- lm(BPSys1 ~ Gender, data = nhanes_bp)
tidy_cat  <- tidy(model_cat, conf.int = TRUE)

# MLR model
model_mlr <- lm(BPSys1 ~ BMI + Gender, data = nhanes_bp)
tidy_mlr  <- tidy(model_mlr, conf.int = TRUE)

# Inline values — MLR
b_bmi_adj  <- lamisc::fmt_num(tidy_mlr$estimate[2], accuracy = 0.01)
b_male_adj <- lamisc::fmt_num(tidy_mlr$estimate[3], accuracy = 0.01)
b_bmi_slr  <- lamisc::fmt_num(tidy(model_slr)$estimate[2], accuracy = 0.01)

r2_slr  <- lamisc::fmt_num(glance(model_slr)$r.squared,  accuracy = 0.001)
r2_mlr  <- lamisc::fmt_num(glance(model_mlr)$r.squared,  accuracy = 0.001)
r2a_mlr <- lamisc::fmt_num(glance(model_mlr)$adj.r.squared, accuracy = 0.001)

# Age group model
model_age <- lm(BPSys1 ~ Age_group, data = nhanes_bp)
tidy_age  <- tidy(model_age, conf.int = TRUE)

n_bp <- nrow(nhanes_bp)
```

# From SLR to Multiple Regression

## Where we left off

\

::::: columns
::: {.column width="55%"}

In the SLR lecture we asked:

> **Does BMI predict systolic blood pressure?**

\

We fit:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 \cdot \text{BMI}$$

\

And found: for every 1 kg/m² increase in BMI, systolic BP is expected to increase by `r b_bmi_slr` mmHg on average.

\

But real biology is messier. **Multiple variables** are usually at play — age, sex, medications, comorbidities...

:::

::: {.column width="45%"}

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 6.0
#| fig-align: center

ggplot(nhanes_bp, aes(x = BMI, y = BPSys1)) +
  geom_point(alpha = 0.35, color = "#0072B2", size = 1.5) +
  geom_smooth(method = "lm", color = "#D55E00", se = FALSE) +
  labs(title = "SLR: BP ~ BMI",
       subtitle = glue("NHANES sample, n = {n_bp}"),
       x = "BMI (kg/m²)", y = "Systolic BP (mmHg)")
```

:::
:::::

## Roadmap for Today

\

::::: columns
::: {.column width="50%"}

**Part 1: Categorical Predictors**

- A new kind of explanatory variable
- The binary categorical model
- Connecting to what you already know
- Interpreting output in R
- Changing the reference category

:::

::: {.column width="50%"}

**Part 2: Multiple Linear Regression**

- Adding predictors to the model
- Confounding: why adjustment matters
- Fitting MLR in R with `lm()`
- Interpreting adjusted coefficients
- $R^2$ and adjusted $R^2$
- Unadjusted vs. adjusted: a side-by-side comparison

:::
:::::

# Part 1: Categorical Predictors

## A new question about blood pressure

\

So far, the explanatory variable has always been **continuous** (e.g., BMI, Age).

\

What if the explanatory variable is **categorical**?

\

::: {.callout-note icon="false"}
### New question

**Does mean systolic blood pressure differ between males and females in the NHANES sample?**

- Response $Y$: Systolic BP (continuous)
- Explanatory $X$: Gender (categorical: female / male)
:::

\

You might think: *"I know how to do this — it's a two-sample t-test!"* And you're right. But regression gives us a unified framework that extends naturally to multiple predictors.

## Visualizing BP by gender

\


::: columns
::: {.column width="40%"}

\

```{r}
#| echo: false

# nhanes_bp %>%
#   group_by(Gender) %>%
#   summarise(n = n(),
#             mean_bp = mean(BPSys1),
#             sd_bp   = sd(BPSys1)) %>%
#   gt() %>%
#   fmt_number(columns = c(mean_bp, sd_bp), decimals = 1) %>%
#   cols_label(Gender = "Gender", n = "n",
#              mean_bp = "Mean BP", sd_bp = "SD") %>%
#   tab_options(table.font.size = px(28))

nhanes_bp %>%
  group_by(Gender) %>%
  summarise(n = n(),
            mean_bp = mean(BPSys1),
            sd_bp   = sd(BPSys1)) %>%
  mutate(dplyr::across(.cols = c(mean_bp, sd_bp), 
                       .fns = ~ lamisc::fmt_num(x = ., accuracy = 0.1))) %>% 
  knitr::kable(col.names = c("Gender", "n", "Mean BP", "SD"))
```

\

Males have a slightly higher mean BP than females in this sample.

:::
::: {.column width="60%"}

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 6.0
#| fig-align: center

ggplot(nhanes_bp, aes(x = Gender, y = BPSys1, fill = Gender)) +
  geom_boxplot(alpha = 0.7, width = 0.4) +
  scale_fill_manual(values = c("female" = "#0072B2", "male" = "#E69F00"),
                    guide = "none") +
  labs(title = "Systolic BP by gender",
       subtitle = glue("NHANES sample, n = {n_bp}"),
       x = NULL, y = "Systolic BP (mmHg)")
```

:::
:::

## The regression model with a binary categorical predictor

\

For a binary categorical variable, `R` creates an **indicator variable** (0/1) automatically:

$$\text{Gender}_{\text{male}} = \begin{cases} 1 & \text{if male} \\ 0 & \text{if female} \end{cases}$$

\

The model becomes:

$$E(\text{BP}) = \beta_0 + \beta_1 \cdot \text{Gender}_{\text{male}}$$

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Intercept $\hat{\beta}_0$
The **estimated mean BP for females** — the group where $\text{Gender}_{\text{male}} = 0$.

This is the **reference category**.
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Slope $\hat{\beta}_1$
The **estimated difference in mean BP** comparing males to females.

$\hat{\beta}_1 = \overline{Y}_{\text{male}} - \overline{Y}_{\text{female}}$
:::

:::
:::::

## Running the model in R

\

```{r}
model_cat <- lm(BPSys1 ~ Gender, data = nhanes_bp)

broom::tidy(model_cat, conf.int = TRUE)
```

\

::: {.callout-important icon="false"}
### Interpretation

- **Intercept** (`r lamisc::fmt_num(tidy_cat$estimate[1], accuracy = 0.1)` mmHg): The estimated mean systolic BP for **females**.
- **Gendermale** (`r lamisc::fmt_num(tidy_cat$estimate[2], accuracy = 0.1)` mmHg): Males have, on average, a systolic BP that is `r lamisc::fmt_num(tidy_cat$estimate[2], accuracy = 0.1)` mmHg **higher** than females (95% CI: `r lamisc::fmt_num(tidy_cat$conf.low[2], accuracy = 0.1)` to `r lamisc::fmt_num(tidy_cat$conf.high[2], accuracy = 0.1)` mmHg).

The p-value (`r lamisc::fmt_pvl(tidy_cat$p.value[2])`) indicates this difference is `r dplyr::if_else(tidy_cat$p.value[2] < 0.05, "", "not ")`statistically significant.
:::

## This is a t-test in disguise

\

::: {.callout-tip icon="false"}
### Connecting to what you already know

Running `lm(BPSys1 ~ Gender)` with a binary predictor gives you the **same p-value** as a two-sample t-test (assuming equal variances).

Regression is the more general framework — it lets us add more predictors, which a t-test cannot do.
:::

\

```{r}
# Two-sample t-test for comparison
t.test(BPSys1 ~ Gender, data = nhanes_bp, var.equal = TRUE)
```

## Changing the reference category

\

R chooses the reference category **alphabetically** — "female" comes before "male", so female is the reference.

\

To change it, use `relevel()`:

\

::: {style="font-size: 0.85em;"}
```{r}
nhanes_bp_male_ref <- nhanes_bp %>%
  mutate(Gender = fct_relevel(Gender, "male"))  # base R equivalent: relevel(Gender, ref = "male")

lm(BPSys1 ~ Gender, data = nhanes_bp_male_ref) %>%
  tidy(conf.int = TRUE)
```
:::

\

::: {.callout-note icon="false"}
The **estimate changes sign** but the **magnitude and p-value are identical** — we're just asking the question from the other direction. Use whichever reference group makes the most scientific sense.
:::

# Part 2: Multiple Linear Regression

## Adding more predictors

\

We have two questions we'd like to answer simultaneously:

- Is BMI associated with BP?
- Is gender associated with BP?

\

**And**: we know BMI and gender are themselves related — males tend to have different BMI distributions than females. This creates a problem called **confounding**.

\

The solution is to put both predictors in the same model:

$$E(\text{BP}) = \beta_0 + \beta_1 \cdot \text{BMI} + \beta_2 \cdot \text{Gender}_{\text{male}}$$

\

::: {.callout-note icon="false"}
### Multiple linear regression (MLR)
A regression model with **two or more** explanatory variables. Each coefficient describes the association between that predictor and the response, **holding the other predictors constant**.
:::

## What is confounding?

\

::::: columns
::: {.column width="55%"}

A **confounder** is a variable that is:

1. Associated with the **explanatory variable** of interest
2. Associated with the **response variable**
3. Not on the causal pathway between them

\

**Example here:** Gender is associated with both BMI and BP.

If we only look at BP ~ BMI, some of the apparent "BMI effect" is really a "gender effect" in disguise.

\

::: {.callout-important icon="false"}
### Adjusting for confounding
Including the confounder in the model lets us estimate the association between BMI and BP **among people of the same gender** — a fairer comparison.
:::

:::

::: {.column width="45%"}

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 6.0
#| fig-align: center

ggplot(nhanes_bp, aes(x = BMI, y = BPSys1, color = Gender)) +
  geom_point(alpha = 0.35, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("female" = "#0072B2", "male" = "#E69F00")) +
  labs(title = "BP vs BMI, stratified by gender",
       x = "BMI (kg/m²)", y = "Systolic BP (mmHg)",
       color = NULL) +
  theme(legend.position = "top")
```

:::
:::::

## Fitting MLR in R

\

The syntax extends naturally from SLR — just add predictors with `+`:

\

```{r}
model_mlr <- lm(BPSys1 ~ BMI + Gender, data = nhanes_bp)

tidy(model_mlr, conf.int = TRUE)
```

\

::: {.callout-note icon="false"}
`R` handles the categorical variable automatically. You do not need to create indicator variables by hand — just make sure `Gender` is a `factor`.
:::

## Interpreting the adjusted coefficients

\

$$\widehat{\text{BP}} = \hat{\beta}_0 + \hat{\beta}_1 \cdot \text{BMI} + \hat{\beta}_2 \cdot \text{Gender}_{\text{male}}$$

\

::: {style="font-size: 0.85em;"}
```{r}
tidy(model_mlr, conf.int = TRUE)
```
:::

::: {.callout-important icon="false"}
### Key rule: always say "holding the other variables constant"

- **BMI** (`r b_bmi_adj` mmHg): For every 1 kg/m² increase in BMI, systolic BP is expected to increase by `r b_bmi_adj` mmHg on average, **holding gender constant**.

- **Gendermale** (`r b_male_adj` mmHg): Males have, on average, a systolic BP that is `r b_male_adj` mmHg higher than females, **holding BMI constant** (i.e., comparing males and females at the same BMI).

- **Intercept**: Expected mean BP for a **female** with BMI = 0. (Not a meaningful prediction — extrapolation.)
:::

## Unadjusted vs. adjusted: what changed?

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Unadjusted (SLR): BP ~ BMI

\

```{r}
#| echo: false

# tidy(model_slr, conf.int = TRUE) %>%
#   select(term, estimate, conf.low, conf.high, p.value) %>%
#   gt() %>%
#   fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) %>%
#   fmt(columns = p.value, fns = lamisc::fmt_pvl) %>%
#   cols_label(term = "Term", estimate = "Est.",
#              conf.low = "Lower", conf.high = "Upper",
#              p.value = "p-value") %>%
#   tab_options(table.font.size = px(26))


tidy(model_slr, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(dplyr::across(.cols = c(estimate, conf.low, conf.high), 
                       .fns = ~ lamisc::fmt_num(x = ., accuracy = 0.01)), 
         p.value = lamisc::fmt_pvl(p.value)) %>% 
  knitr::kable(col.names = c("Term", "Est.", "Lower", "Upper", "p-value"), 
               align = "lr")
```

\

BMI slope: **`r b_bmi_slr`** mmHg per kg/m²

:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Adjusted (MLR): BP ~ BMI + Gender

\

```{r}
#| echo: false

# tidy(model_mlr, conf.int = TRUE) %>%
#   select(term, estimate, conf.low, conf.high, p.value) %>%
#   gt() %>%
#   fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) %>%
#   fmt(columns = p.value, fns = lamisc::fmt_pvl) %>%
#   cols_label(term = "Term", estimate = "Est.",
#              conf.low = "Lower", conf.high = "Upper",
#              p.value = "p-value") %>%
#   tab_options(table.font.size = px(26))

tidy(model_mlr, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(dplyr::across(.cols = c(estimate, conf.low, conf.high), 
                       .fns = ~ lamisc::fmt_num(x = ., accuracy = 0.01)), 
         p.value = lamisc::fmt_pvl(p.value)) %>% 
  knitr::kable(col.names = c("Term", "Est.", "Lower", "Upper", "p-value"), 
               align = "lr")
```

BMI slope: **`r b_bmi_adj`** mmHg per kg/m²

:::

:::
:::::

\

::: {.callout-tip icon="false"}
The BMI coefficient changed slightly after adjusting for gender — a sign that gender was acting as a mild confounder. In some datasets, adjustment can dramatically change an estimate (or even flip its direction).
:::

## $R^2$ and adjusted $R^2$

\

You've seen $R^2$ as a measure of model fit. In MLR there's a wrinkle:

\

::::: columns
::: {.column width="55%"}

::: {.callout-note icon="false"}
### $R^2$ always increases when you add predictors
Even if a new predictor is pure noise, $R^2$ can only stay the same or go up — making it a poor tool for comparing models with different numbers of predictors.
:::

\

**Adjusted $R^2$** penalizes for the number of predictors:

$$R^2_{\text{adj}} = 1 - \frac{(1-R^2)(n-1)}{n - p - 1}$$

where $p$ = number of predictors. A new predictor only improves adjusted $R^2$ if it adds more than it costs.

:::

::: {.column width="5%"}
:::

::: {.column width="40%"}
```{r}
glance(model_mlr) %>% 
  select(r.squared, 
         adj.r.squared, 
         sigma, 
         nobs)
```

\

In our model: $R^2$ = `r r2_mlr`, adjusted $R^2$ = `r r2a_mlr`.

:::
:::::

## The MLR workflow

\

::::: columns
::: {.column width="45%"}

**1. Visualize**

```r
ggplot(data, aes(x = X1, y = Y, color = X2)) +
  geom_point() +
  geom_smooth(method = "lm")
```

\

**2. Ensure categorical vars are factors**

```r
data <- data %>%
  mutate(X2 = factor(X2))
```

\

**3. Fit**

```r
model <- lm(Y ~ X1 + X2, data = data)
```

:::

::: {.column width="3%"}
:::

::: {.column width="52%"}

**4. Coefficients + inference**

```r
tidy(model, conf.int = TRUE)
```

\

**5. Model fit**

```r
glance(model)   # R², adj. R², sigma
```

\

**6. Change reference category (if needed)**

::: {style="font-size: 0.85em;"}
```r
data <- data %>%
  mutate(X2 = foracts::fct_relevel(X2, "your_ref"))
```
:::

\

**7. Interpret everything as "adjusted"**

Every coefficient describes the association with $Y$ **holding all other predictors constant**.

:::
:::::

## Categorical predictors with 3+ levels

\

What if a categorical variable has **more than two levels**?

\

R creates **one coefficient per level, each compared to the same reference group** — the reference category is absorbed into the intercept and doesn't get its own row.

\

Let's create age groups from the continuous `Age` variable and fit the model (next slide):

\
```{r}
nhanes_bp <- nhanes_bp %>%
  mutate(Age_group = cut(Age,
                         breaks = c(0, 39, 59, Inf),
                         labels = c("Under 40", "40–59", "60+")))

nhanes_bp %>% 
  janitor::tabyl(Age_group)

```

## Interpreting the output

\

::: {style="font-size: 0.85em;"}
```{r}
model_age <- lm(BPSys1 ~ Age_group, data = nhanes_bp)

tidy(model_age, conf.int = TRUE)
```
:::

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### Reference category: "Under 40"
The intercept (`r lamisc::fmt_num(tidy_age$estimate[1], accuracy = 0.1)` mmHg) is the estimated mean BP for participants **under 40** — the group that doesn't get its own row.
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-important icon="false"}
### Interpretation template
- **40–59**: mean BP is `r lamisc::fmt_num(tidy_age$estimate[2], accuracy = 0.1)` mmHg **higher than the under-40 group**
- **60+**: mean BP is `r lamisc::fmt_num(tidy_age$estimate[3], accuracy = 0.1)` mmHg **higher than the under-40 group**

Every coefficient is *always* relative to the reference — never relative to each other.
:::

:::
:::::

## What we haven't covered (and where to go)

\

MLR is a huge topic. Today's lecture gives you the foundation; there's much more in the textbook (Chapter 7) and beyond:

\

::::: columns
::: {.column width="48%"}

::: {.callout-note icon="false"}
### In your textbook (Ch. 7)
- Model selection strategies (§7.8)
- Interaction terms (§7.7)
- Connection to ANOVA (§7.9)
:::

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

::: {.callout-note icon="false"}
### Where regression leads
- **Logistic regression** — binary response (yes/no, alive/dead)
- **Survival analysis** — time-to-event outcomes
- **Mixed models** — correlated observations 
:::

:::
:::::
